---
title: "NSW Road Crash Severity Prediction"
subtitle: "STAT5003  -  Assignment 2 Group 07, Workshop 12"
author: "Sanskar | Robert | Haipeng | Chen | Yihan"
date: "23 May 2026"
output:
  html_document:
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  results = 'hide',
  warning = FALSE,
  message = FALSE
)
```

```{r library, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(kableExtra)
library(caret)
library(nnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(kknn)
library(e1071)
library(themis)
library(pROC)
library(flextable)
library(MLmetrics)
library(Matrix)
library(tibble)
library(naivebayes)
library(cowplot)
library(gridExtra)

# Create plots output folder  -  all ggsave() calls write here for use in Presentation Rmd
dir.create("plots", showWarnings = FALSE)
```

```{css, echo=FALSE}
body { font-family: "Times New Roman", Times, serif; font-size: 10pt; line-height: 1.3; }
h2 { margin-top: 0.4em; margin-bottom: 0.15em; }
h3 { margin-top: 0.3em; margin-bottom: 0.1em; }
h4 { margin-top: 0.25em; margin-bottom: 0.1em; }
p  { margin-top: 0.2em; margin-bottom: 0.2em; }
h1.title { display: none; }
h4.author { display: none; }
h4.date   { display: none; }
```

```{r load-data, message=FALSE, warning=FALSE, results='hide'}
df <- read.csv("nsw_road_crash_merged_2020_2024.csv", check.names = FALSE)
cat("Dimensions:", nrow(df), "rows x", ncol(df), "cols\n")
str(df)
```

------------------------------------------------------------------------

### 1 Problem Framing & Research Question

#### 1.1 Research Question
Given environmental conditions (weather, natural lighting, road surface, speed limit) and traffic unit characteristics (type of traffic unit, manoeuvre, number of units involved), can we predict the severity of a road crash in NSW as a three-class classification problem: Fatal, Injury, or Non-casualty?

#### 1.2 Motivation
NSW recorded 340 road fatalities in 2024, with over two-thirds occurring in regional areas (Transport for NSW, 2024), highlighting the need for data-driven approaches to understanding crash severity. Fatal crashes represent only approximately 1.4% of all crashes, introducing severe class imbalance that must be carefully addressed during modelling and evaluation.

A predictive model for crash severity has direct real-world value across three stakeholder groups:

```{r stakeholder-table, echo=FALSE, results='asis'}
stakeholder_df <- data.frame(
  Stakeholder = c("NSW Traffic Management", "Emergency Response Teams", "Policymakers & Insurers"),
  `Use Case` = c(
    "Targeted speed-limit adjustment; road-design upgrades in high-risk corridors",
    "Early-stage dispatch allocation for ambulance and emergency services",
    "Evidence-based road safety policy; risk-adjusted insurance pricing"
  ),
  `How the model helps` = c(
    "Identifies which environmental and traffic-unit combinations are associated with severe outcomes",
    "Predicts severity from observable conditions at incident onset before full details are available",
    "Quantifies which vehicle types and conditions elevate severity risk for policy and pricing models"
  ),
  check.names = FALSE
)
knitr::kable(stakeholder_df, caption = "Table 1: Stakeholder Value") |>
  kableExtra::kable_styling(full_width = TRUE,
                            bootstrap_options = c("condensed"), font_size = 11) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::column_spec(1, bold = TRUE, width = "22%") |>
  kableExtra::column_spec(2, width = "35%") |>
  kableExtra::column_spec(3, width = "43%")
```

------------------------------------------------------------------------

### 2 Dataset & Justification

#### 2.1 Dataset Overview

Two source tables from Transport for NSW were merged via left join on `Crash ID`: **CRASH.xlsx** provides crash-level context (time, location, environment, speed limit) across 92,189 unique incidents, while **TRAFFIC UNIT.xlsx** contributes per-vehicle attributes (unit type, manoeuvre, controlled flag) across 170,962 traffic unit records. The merged dataset therefore contains one row per traffic unit, enabling vehicle-specific features to enrich crash severity prediction alongside crash-level environmental factors. The three-class target (Fatal, Injury, and Non-casualty) is severely imbalanced, with Fatal crashes representing only ~1.4% of all records, making this imbalance the central modelling challenge addressed throughout Sections 4–7.

```{r dataset-overview, echo=FALSE, message=FALSE, warning=FALSE}
overview <- data.frame(
  Item = c(
    "Time Period",
    "Data Sources",
    "Merge Key",
    "Total Records",
    "Total Variables",
    "Target Variable",
    "Fatal",
    "Injury",
    "Non-casualty"
  ),
  Detail = c(
    "2020 - 2024",
    "CRASH.xlsx + TRAFFIC UNIT.xlsx (TfNSW via data.gov.au)",
    "Crash ID (left join)",
    format(nrow(df), big.mark = ","),
    as.character(ncol(df)),
    "Degree of Crash (3-class classification)",
    paste0(format(sum(df$`Degree of crash` == "Fatal"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Fatal") * 100, 1), "%)"),
    paste0(format(sum(df$`Degree of crash` == "Injury"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Injury") * 100, 1), "%)"),
    paste0(format(sum(df$`Degree of crash` == "Non-casualty (towaway)"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Non-casualty (towaway)") * 100, 1), "%)")
  )
)

knitr::kable(overview, col.names = c("Item", "Detail"),
             caption = "Table 2: Dataset Overview")
```

#### 2.2 Why this dataset?

```{r dataset-justification, echo=FALSE, results='asis'}
justif_df <- data.frame(
  Criterion = c("Large", "Integrated", "Messy", "Multi-class classification"),
  Detail = c(
    "170,962 records spanning 5 years (2020-2024)",
    "Two source tables joined on Crash ID - crash-level attributes merged with per-traffic-unit attributes",
    "Missing values, outlier GPS coordinates, rare factor levels, class imbalance (~1.4% Fatal)",
    "Three-class target: Fatal / Injury / Non-casualty - directly matching real-world severity categories"
  ),
  `Relevance to A2` = c(
    "Sufficient volume to support train/test/holdout splits and 5-fold CV without data starvation",
    "Requires careful join strategy and leakage-free preprocessing to avoid contaminating predictions",
    "Demonstrates full preprocessing pipeline: imputation, encoding, outlier handling, imbalance correction",
    "Tests all classifiers on a realistic, imbalanced multi-class problem with a meaningful minority class"
  ),
  check.names = FALSE
)
knitr::kable(justif_df, caption = "Table 2: Dataset Justification") |>
  kableExtra::kable_styling(full_width = TRUE,
                            bootstrap_options = c("condensed"), font_size = 11) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::column_spec(1, bold = TRUE, width = "20%") |>
  kableExtra::column_spec(2, width = "38%") |>
  kableExtra::column_spec(3, width = "42%")
```

------------------------------------------------------------------------

### 3 EDA & Insights
```{r 311-selected, fig.width=13, fig.height=8, message=FALSE, warning=FALSE}
library(gridExtra)

p1 <- ggplot(df, aes(x = fct_infreq(`Degree of crash`))) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3) +
    labs(title = "Distribution of Crash Severity", x = "Severity", y = "Count") +
    theme_minimal() + theme(legend.position = "none")

p2 <- ggplot(df, aes(x = `Speed limit`)) +
  geom_bar(fill = "tomato") +
  labs(title = "Crash Distribution by Speed Limit", x = "Speed Limit", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 <- ggplot(df, aes(x = fct_infreq(`TU type group`))) +
    geom_bar(fill = "steelblue") +
    labs(title = "Crash Distribution by Traffic Unit Type", x = "Traffic Unit Type", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4 <- ggplot(df, aes(x = fct_infreq(Urbanisation))) +
  geom_bar(fill = "purple") +
  labs(title = "Crash Distribution by Urbanisation", x = "Urbanisation", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Bivariate: crash severity proportion by speed limit
df_speed_sev <- df %>%
  filter(!is.na(`Speed limit`), !is.na(`Degree of crash`)) %>%
  mutate(`Speed limit` = factor(`Speed limit`,
    levels = c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h",
               "60 km/h","70 km/h","80 km/h","90 km/h","100 km/h","110 km/h","Unknown")))

p5 <- ggplot(df_speed_sev, aes(x = `Speed limit`, fill = `Degree of crash`)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Severity Proportion by Speed Limit",
       x = "Speed Limit", y = "Proportion", fill = "Severity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.key.size = unit(0.4, "cm"), legend.text = element_text(size = 8))

# Bivariate: fatal rate by TU type
tu_fatal_eda <- df %>%
  filter(!is.na(`TU type group`), !is.na(`Degree of crash`)) %>%
  group_by(`TU type group`) %>%
  summarise(total = n(), fatal_rate = mean(`Degree of crash` == "Fatal"), .groups = "drop") %>%
  filter(total >= 100) %>%
  arrange(desc(fatal_rate))

p6 <- ggplot(tu_fatal_eda, aes(x = reorder(`TU type group`, fatal_rate), y = fatal_rate)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(title = "Fatal Rate by Traffic Unit Type",
       x = NULL, y = "Fatal %") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3, nrow = 2)
```

- **Crash Severity:** Fatal crashes represent only 1.4% of records - this severe class imbalance means a naive classifier predicting "Injury" for everything achieves ~67% accuracy while missing every Fatal crash, making addressing imbalance the single most important preprocessing decision for this problem.
- **Speed Limit:** Crashes cluster heavily in 50-60 km/h urban zones, reflecting traffic volume rather than per-trip risk. However, high-speed zones (90-110 km/h) are likely disproportionately linked to Fatal outcomes, making speed limit a key candidate predictor of severity.
- **Traffic Unit Type:** Car/car derivatives dominate volume (~114,477) but pedestrians and motorcycles - though rare - are strongly associated with severe and fatal outcomes. Their low frequency relative to cars reinforces the minority-class challenge and motivates the inclusion of has_pedestrian and has_motorcycle engineered flags.
- **Urbanisation:** Sydney metro generates the most crashes (98,975) but Country non-urban regions show higher Fatal proportions despite lower volume - consistent with higher speed limits, longer emergency response times, and reduced road quality outside urban centres. This geographic signal directly motivates including LGA and Urbanisation as predictors.
- **Severity by Speed Limit (bivariate):** Fatal proportion generally increases from low-speed urban zones (≤50 km/h) to high-speed rural zones (90-110 km/h), confirming that raw speed limit carries ordinal severity signal. This non-linear step pattern directly motivates grouping speed limits into `speed_band` (low_50 / mid_60_80 / high_90_100 / very_high_110) rather than treating the variable as unordered categorical in preprocessing (Step 4.4).
- **Fatal Rate by TU Type (bivariate):** Pedestrians have fatal rates up to 10× higher than car occupants, and motorcyclists around 5-6× despite contributing a small share of overall volume. This confirms that TU type is among the strongest severity predictors and directly justifies engineering `has_pedestrian` and `has_motorcycle` binary flags (Step 4.1) to make these high-risk unit types explicitly visible to all classifiers.

------------------------------------------------------------------------

### 4 Data Preprocessing

Steps 4.1–4.6 (integration, feature removal, imputation, feature engineering, outlier handling, and encoding) are applied to the **full dataset before** the train/test split. Downsampling to `df_train_bal` is applied to the **training partition only**, after the split. SMOTE oversampling (LR and NB) is applied **within cross-validation folds** to prevent resampling leakage. Feature scaling for KNN is applied to the full training set via caret's `preProcess`, as KNN uses a direct fit with no CV (James et al., 2021).

```{r preprocessing-all, echo=TRUE, results='hide'}
# 4.1 Data Integration
# Each row is one traffic unit - summarise binary flags per crash, then join back
tu_flags <- df %>%
  group_by(`Crash ID`) %>%
  summarise(
    has_motorcycle    = as.integer(any(`TU type group` == "Motorcycle")),
    has_heavy_vehicle = as.integer(any(`TU type group` %in% c("Articulated truck", "Heavy rigid truck", "Bus"))),
    has_pedestrian    = as.integer(any(`TU type group` == "Pedestrian")),
    has_cyclist       = as.integer(any(`TU type group` == "Pedal cycle")),
    tu_count          = n(),
    .groups = "drop"
  )
df <- df %>% left_join(tu_flags, by = "Crash ID")

# 4.2 Feature Removal
leakage_cols <- c("No. killed", "No. seriously injured",
                  "No. moderately injured", "No. minor-other injured")
redundant_target_cols <- c("Degree of crash - detailed")
id_cols <- c("Crash ID", "Traffic unit ID", "Street of crash",
             "Street of travel", "Town", "Identifying feature",
             "Identifying feature type", "Conurbation 1")
high_na_cols <- c("Primary hazardous feature", "Primary temporary feature",
                  "Object hit 2", "DCA supplement", "Object hit 1",
                  "Primary permanent feature")
redundant_cols <- c("Route no.", "RUM - code", "RUM - description",
                    "DCA - code", "DCA - description")
drop_cols <- unique(c(leakage_cols, redundant_target_cols, id_cols, high_na_cols, redundant_cols))
df_clean  <- df %>% select(-any_of(drop_cols))

# 4.3 Missing Value Imputation
df_clean <- df_clean %>%
  mutate(
    `Other TU type` = ifelse(is.na(`Other TU type`), "Unknown", `Other TU type`),
    across(where(is.character), ~ replace_na(., "Unknown")),
    across(where(is.numeric),   ~ replace_na(., median(., na.rm = TRUE)))
  )

# 4.4 Feature Engineering
df_clean <- df_clean %>%
  mutate(
    crash_hour   = as.integer(sub(":.*", "", `Two-hour intervals`)),
    is_weekend   = as.integer(`Day of week of crash` %in% c("Saturday", "Sunday")),
    is_rush_hour = as.integer(crash_hour %in% c(7, 8, 9, 16, 17, 18)),
    is_night     = as.integer(`Natural lighting` %in% c("Darkness", "Dawn", "Dusk")),
    is_wet       = as.integer(`Surface condition` %in% c("Wet", "Snow or ice")),
    speed_band   = case_when(
      `Speed limit` %in% c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h") ~ "low_50",
      `Speed limit` %in% c("60 km/h","70 km/h","80 km/h")                     ~ "mid_60_80",
      `Speed limit` %in% c("90 km/h","100 km/h")                              ~ "high_90_100",
      `Speed limit` == "110 km/h"                                              ~ "very_high_110",
      TRUE                                                                     ~ "unknown"
    )
  )

# 4.5 Outlier Handling
p99_dist <- quantile(df_clean$Distance, 0.99, na.rm = TRUE)
df_clean <- df_clean %>%
  mutate(Distance = pmin(Distance, p99_dist)) %>%
  filter(is.na(Latitude)  | (Latitude  >= -37.5 & Latitude  <= -28.2),
         is.na(Longitude) | (Longitude >= 141.0  & Longitude <= 153.6))

# 4.6 Encoding
lga_freq  <- prop.table(table(df_clean$LGA))
rare_lgas <- names(lga_freq[lga_freq < 0.01])
df_clean  <- df_clean %>%
  mutate(LGA = ifelse(LGA %in% rare_lgas, "Other", LGA))
df_model <- df_clean %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(`Degree of crash` = factor(`Degree of crash`,
           levels = c("Fatal", "Injury", "Non-casualty (towaway)"),
           labels = c("Fatal", "Injury", "Non_casualty")))

# 4.7 Class Imbalance - 2024 sealed as temporal holdout; undersample Injury/Non-casualty to 15k each
df_holdout_2024  <- df_model %>% filter(`Reporting year` == 2024)
df_model_2020_23 <- df_model %>% filter(`Reporting year` != 2024)
set.seed(42)
train_idx <- createDataPartition(df_model_2020_23$`Degree of crash`, p = 0.9, list = FALSE)
df_train  <- df_model_2020_23[ train_idx, ]
df_test   <- df_model_2020_23[-train_idx, ]

fatal_rows   <- df_train %>% filter(`Degree of crash` == "Fatal")
injury_rows  <- df_train %>% filter(`Degree of crash` == "Injury") %>% slice_sample(n = 15000)
noncas_rows  <- df_train %>% filter(`Degree of crash` == "Non_casualty") %>% slice_sample(n = 15000)
df_train_bal <- bind_rows(fatal_rows, injury_rows, noncas_rows) %>% slice_sample(prop = 1)

# Inverse frequency weights
class_weights <- 1 / prop.table(table(df_train_bal$`Degree of crash`))
class_weights <- class_weights / sum(class_weights)

# CV strategy - not all models use caret's trainControl; each uses the method best suited to its library:
#   cv_ctrl_smote : LR (baseline + tuned), NB (baseline + tuned) - SMOTE applied inside each CV fold
#   rpart xval    : DT (baseline + tuned) - rpart's built-in k-fold CP pruning via xval parameter
#   OOB error     : RF (baseline + tuned) - randomForest's out-of-bag estimate is an unbiased CV equivalent
#   Early stopping: XGBoost - xgb.train watchlist monitors a held-out validation fold each grid iteration
#   method="none" : KNN - no CV needed; k fixed by sqrt(p) rule (Cover & Hart, 1967); see Section 6.5
cv_ctrl_smote <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final",
  sampling = "smote"
)

# 4.8 Train/Test Split
cat("Train (balanced):", nrow(df_train_bal),
    "| Test:", nrow(df_test),
    "| Holdout (2024):", nrow(df_holdout_2024), "\n")
cat("df_train_bal class distribution:\n")
print(prop.table(table(df_train_bal$`Degree of crash`)))
```

```{r preprocessing-table, echo=FALSE, results='asis'}
prep_df <- data.frame(
  Step = c("4.1 Integration", "4.2 Feature removal", "4.3 Imputation",
           "4.4 Feature engineering", "4.5 Outlier handling", "4.6 Encoding",
           "4.7 Class imbalance", "4.8 Train/test split"),
  Action = c(
    "Aggregate traffic-unit rows into per-crash binary flags (has_motorcycle, has_pedestrian, etc.)",
    "Drop leakage cols (injury counts), identifiers, free-text fields, cols >50% missing",
    "Character NAs to 'Unknown'; numeric NAs to column median (median robust to skew)",
    "Derive is_weekend, is_rush_hour, is_night, is_wet, speed_band from existing columns",
    "Winsorise Distance at 99th pct; filter GPS outside NSW bounds",
    "Pool LGA levels <1% frequency into 'Other'; encode all characters as factors",
    "Retain all ~770 Fatal rows; downsample Injury & Non-casualty to 15,000 each, giving ~31,500 rows (ratio ~1:10:10). SMOTE oversampling (LR, NB) is applied within CV folds. KNN scaling is applied to the full training set (no CV, direct fit).",
    "90/10 stratified split on 2020-2023; 2024 sealed as temporal holdout"
  ),
  Justification = c(
    "Left join creates one row per traffic unit; aggregation prevents duplicating crash-level features",
    "Injury count columns leak the target; identifiers add no predictive signal",
    "Median chosen over mean - Distance and other numerics are right-skewed",
    "Time-of-day and surface context not directly available as numeric features",
    "A few extreme distances and erroneous coordinates would distort distance-based models",
    "Reduces LGA cardinality from ~150 to manageable levels before OHE",
    "1:47:23 ratio too extreme for Fatal detection; equal majority-class caps avoid secondary 2:1 imbalance",
    "Temporal split prevents future data leaking into training; stratification preserves Fatal proportion"
  )
)
knitr::kable(prep_df, col.names = c("Step", "Action", "Justification"),
             caption = "Table 3: Preprocessing Pipeline") |>
  kableExtra::kable_styling(full_width = TRUE,
                            bootstrap_options = c("condensed"), font_size = 11) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::column_spec(1, bold = TRUE, width = "15%") |>
  kableExtra::column_spec(2, width = "42%") |>
  kableExtra::column_spec(3, width = "43%")
```

**Imbalance strategy - random downsampling:**

- **The raw 2020-2023 training split has a Fatal**:Injury:Non-casualty ratio of approximately 1:47:23. Training directly on this distribution causes models to optimise for the majority classes and almost entirely ignore the Fatal minority, producing near-zero Fatal Recall on the test set.
- **Random downsampling** (retaining all ~770 Fatal rows, capping Injury and Non-casualty at 15,000 each) was chosen as the primary rebalancing strategy. The resulting ~31,500-row balanced frame (`df_train_bal`) reduces the ratio to approximately 1:10:10 - still imbalanced in favour of majority classes, but a significant improvement over 1:47.
- **Why not SMOTE?** Synthetic oversampling (Chawla et al., 2002) requires a sufficiently dense neighbourhood around minority-class points to generate plausible synthetic examples. With only ~770 Fatal rows spanning a wide range of road types, lighting conditions, and geographic locations, the neighbourhood structure is too sparse for reliable synthesis - SMOTE was applied inside CV folds for LR and NB only as a secondary measure.
- **Why not train on the full dataset?** Initial experiments with the unbalanced full training set (~140,000 rows) resulted in prohibitively long compute times per CV fold (>45 minutes per knit), making iterative experimentation impractical. Downsampling to ~31,500 rows reduced per-fold training time to a tractable level while retaining sufficient sample diversity.
- **Equal majority-class caps (15,000 each):** Proportional downsampling would preserve a ~2:1 Injury/Non-casualty ratio within the training set, causing uneven learning of the Injury/Non-casualty boundary. Equal caps ensure both majority classes are represented with comparable fidelity.
- **Known limitations:** Downsampling discards approximately 80% of available Injury and Non-casualty data, potentially losing rare within-class combinations that could improve boundary precision. The artificially balanced training distribution also diverges from the natural ~1.4% Fatal prevalence in deployment, which is a key driver of the generalisation drop observed in the 2024 holdout (Section 8).
- **Inverse-frequency class weights** were additionally applied to the Decision Tree to further penalise Fatal misclassifications within the downsampled training set.

<!-- ------------------------------------------------------------------------ -->

### 5 Model Fitting

Six classifiers were fitted to cover a range of complexity and interpretability - from a parametric baseline (Logistic Regression) through tree ensembles (Decision Tree, Random Forest, XGBoost) to distance- and probability-based models (KNN, Naive Bayes). All models train on the same `df_train_bal` (~31,500 rows) and are evaluated on the same held-out test set, so cross-model differences reflect genuine algorithmic variation rather than data preparation differences.

```{r model-summary-table, echo=FALSE, results='asis'}
model_sum <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest", "XGBoost", "KNN", "Naive Bayes"),
  Type = c("Parametric linear", "Non-parametric tree", "Ensemble (bagging)", "Ensemble (boosting)", "Instance-based", "Probabilistic"),
  `Key Assumption` = c(
    "Linear log-odds boundary between classes",
    "Axis-aligned recursive splits",
    "Decorrelated trees reduce variance",
    "Additive model minimises log-loss sequentially",
    "Similar instances share the same class",
    "Features conditionally independent given class"
  ),
  `Imbalance handling` = c(
    "SMOTE inside CV folds",
    "Inverse-frequency class weights",
    "None (df_train_bal balanced; OOB controls overfitting)",
    "Inverse-frequency observation weights",
    "None (df_train_bal already balanced)",
    "SMOTE inside CV folds"
  ),
  check.names = FALSE
)
knitr::kable(model_sum, caption = "Table 4: Model Summary") |>
  kableExtra::kable_styling(full_width = TRUE,
                            bootstrap_options = c("condensed"), font_size = 11) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::column_spec(1, bold = TRUE, width = "17%") |>
  kableExtra::column_spec(2, width = "15%") |>
  kableExtra::column_spec(3, width = "35%") |>
  kableExtra::column_spec(4, width = "33%")
```

#### 5.1 Logistic Regression

```{r model-lr, echo=TRUE, results='hide', fig.show='hide'}
# 5.1 Logistic Regression - baseline fit; class imbalance handled via SMOTE inside CV folds
train_lr <- df_train_bal %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

test_lr <- df_test %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

model_lr <- train(
  Degree.of.crash ~ .,
  data = train_lr,
  method = "multinom",
  trControl = cv_ctrl_smote,
  preProcess = "nzv",
  metric = "Mean_F1",
  trace = FALSE,
  MaxNWts = 20000
)

print(model_lr$results)

pred_lr <- predict(model_lr, newdata = test_lr)
cm_lr   <- confusionMatrix(pred_lr, test_lr$Degree.of.crash)

fatal_recall_lr <- as.numeric(cm_lr$byClass["Class: Fatal", "Recall"])
f1_lr_vec <- setNames(cm_lr$byClass[, "F1"], gsub("Class: ", "", rownames(cm_lr$byClass)))
test_props_lr <- prop.table(table(test_lr$Degree.of.crash))
macro_f1_lr <- mean(f1_lr_vec, na.rm = TRUE)
weighted_f1_lr <- weighted.mean(f1_lr_vec[names(test_props_lr)], test_props_lr, na.rm = TRUE)
kappa_lr <- as.numeric(cm_lr$overall["Kappa"])

cat(sprintf(
  "   Logistic Regression  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_lr, fatal_recall_lr, weighted_f1_lr, kappa_lr
))

# Confusion matrix heatmap
p_lr_cm <- as.data.frame(cm_lr$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Logistic Regression - Confusion Matrix",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_lr, 3),
                      "  |  Macro-F1: ", round(macro_f1_lr, 3)),
    x = "Actual", y = "Predicted"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 10))
print(p_lr_cm)
ggsave("plots/5_1_logistic_regression_confusion_matrix.png", p_lr_cm, width = 7, height = 5, dpi = 150)

# Coefficient plot  -  top 15 absolute coefficients across non-reference classes
coef_lr <- coef(model_lr$finalModel)

if (is.null(dim(coef_lr))) {
  coef_lr <- matrix(coef_lr, nrow = 1)
  rownames(coef_lr) <- "Non-reference class"
}

coef_lr_df <- as.data.frame(as.table(coef_lr))
if (ncol(coef_lr_df) == 3) {
  names(coef_lr_df) <- c("Class", "Feature", "Coefficient")
} else if (ncol(coef_lr_df) == 2) {
  names(coef_lr_df) <- c("Feature", "Coefficient")
  coef_lr_df$Class <- "Non-reference class"
  coef_lr_df <- coef_lr_df %>% select(Class, Feature, Coefficient)
}

coef_lr_df <- coef_lr_df %>%
  filter(Feature != "(Intercept)") %>%
  mutate(abs_coef = abs(as.numeric(Coefficient))) %>%
  arrange(desc(abs_coef)) %>%
  slice_head(n = 15)

p_lr_coef <- ggplot(coef_lr_df, aes(x = reorder(Feature, abs_coef), y = as.numeric(Coefficient), fill = Class)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Logistic Regression  -  Largest Coefficients (Top 15)",
    x = NULL,
    y = "Coefficient (log-odds)"
  ) +
  theme_minimal(base_size = 12)
print(p_lr_coef)
ggsave("plots/5_1_logistic_regression_coefficients.png", p_lr_coef, width = 7, height = 5, dpi = 150)
```

#### 5.2 Decision Tree

```{r model-dt, echo=TRUE, results='hide', fig.show='hide'}
# Direct rpart(); OHE via dummyVars; caret formula interface unstable here

# 1. First standardize column names with `make.names` to prevent `dummyVars` from failing to locate columns due to spaces.
train_dt_clean <- df_train_bal %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

test_dt_clean <- df_test %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

# 2. Establish One-Hot encoding
dummy_model <- dummyVars(Degree.of.crash ~ ., data = train_dt_clean)

# 3. Transform the training set
train_dt_x <- as.data.frame(predict(dummy_model, newdata = train_dt_clean))
train_dt <- cbind(train_dt_x, Degree.of.crash = train_dt_clean$Degree.of.crash)

# 4. Transform the test set and concatenate back the target variable
test_dt_x <- as.data.frame(predict(dummy_model, newdata = test_dt_clean))
test_dt <- cbind(test_dt_x, Degree.of.crash = test_dt_clean$Degree.of.crash)

# 5. Align weights
dt_weights <- as.numeric(class_weights[as.character(train_dt$Degree.of.crash)])

# 6. Train the decision tree
model_dt <- rpart(
  Degree.of.crash ~ .,
  data = train_dt,
  method = "class",
  weights = dt_weights,
  control = rpart.control(cp = 0.01, xval = 10)
)

pred_dt <- predict(model_dt, newdata = test_dt, type = "class")
cm_dt   <- confusionMatrix(pred_dt, test_dt$Degree.of.crash)

fatal_recall_dt <- as.numeric(cm_dt$byClass["Class: Fatal", "Recall"])
f1_dt_vec <- setNames(cm_dt$byClass[, "F1"], gsub("Class: ", "", rownames(cm_dt$byClass)))
test_props_dt <- prop.table(table(test_dt$Degree.of.crash))
macro_f1_dt <- mean(f1_dt_vec, na.rm = TRUE)
weighted_f1_dt <- weighted.mean(f1_dt_vec[names(test_props_dt)], test_props_dt, na.rm = TRUE)
kappa_dt <- as.numeric(cm_dt$overall["Kappa"])

cat(sprintf(
  "   Decision Tree  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_dt, fatal_recall_dt, weighted_f1_dt, kappa_dt
))

printcp(model_dt)

# Tree diagram  -  use a simpler, more robust plotting configuration
png("plots/5_2_decision_tree_diagram.png", width = 1200, height = 800, res = 150)
rpart.plot(
  model_dt,
  type = 2,
  extra = 1,
  under = TRUE,
  fallen.leaves = TRUE,
  cex = 0.7,
  main = "Decision Tree Diagram"
)
invisible(dev.off())

rpart.plot(
  model_dt,
  type = 2,
  extra = 1,
  under = TRUE,
  fallen.leaves = TRUE,
  cex = 0.7,
  main = "Decision Tree Diagram"
)

# Confusion matrix heatmap
p_dt_cm <- as.data.frame(cm_dt$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Decision Tree - Confusion Matrix",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_dt, 3),
                      "  |  Macro-F1: ", round(macro_f1_dt, 3)),
    x = "Actual", y = "Predicted"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 10))
print(p_dt_cm)
ggsave("plots/5_2_decision_tree_confusion_matrix.png", p_dt_cm, width = 7, height = 5, dpi = 150)
```

#### 5.3 Random Forest

```{r model-rf, echo=TRUE, results='hide', fig.show='hide'}
# randomForest() direct; OOB used as CV estimate; na.roughfix() for residual NAs
rf_x_train <- na.roughfix(df_train_bal %>%
  select(-`Degree of crash`) %>%
  select(where(~ !is.factor(.) || nlevels(.) <= 53)) %>%
  droplevels())

# Align test factor levels to training  -  unseen levels become NA, then imputed
rf_x_test_raw <- df_test %>% select(all_of(names(rf_x_train)))
for (col in names(rf_x_test_raw)) {
  if (is.factor(rf_x_test_raw[[col]]))
    rf_x_test_raw[[col]] <- factor(rf_x_test_raw[[col]], levels = levels(rf_x_train[[col]]))
}
rf_x_test <- na.roughfix(rf_x_test_raw)

model_rf <- randomForest(
  x          = rf_x_train,
  y          = df_train_bal$`Degree of crash`,
  mtry       = floor(sqrt(ncol(rf_x_train))),
  ntree      = 100,
  importance = TRUE
)

# Test set predictions and metrics
pred_rf         <- predict(model_rf, rf_x_test)
cm_rf           <- confusionMatrix(pred_rf, df_test$`Degree of crash`)
fatal_recall_rf <- cm_rf$byClass["Class: Fatal", "Recall"]
macro_f1_rf     <- mean(cm_rf$byClass[, "F1"],  na.rm = TRUE)
weighted_f1_rf  <- weighted.mean(cm_rf$byClass[, "F1"],
                    prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_rf        <- cm_rf$overall["Kappa"]

cat(sprintf(
  "   Random Forest  -   Test Set Metrics\nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf$err.rate[100, "OOB"], macro_f1_rf, fatal_recall_rf, weighted_f1_rf, kappa_rf
))

# Confusion matrix heatmap
p_rf_cm <- as.data.frame(cm_rf$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(title = "Random Forest - Confusion Matrix",
       subtitle = paste0("Fatal Recall: ", round(fatal_recall_rf, 3),
                         "  |  Macro-F1: ", round(macro_f1_rf, 3)),
       x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 10))
print(p_rf_cm)
ggsave("plots/5_3_random_forest_confusion_matrix.png", p_rf_cm, width = 7, height = 5, dpi = 150)

# Variable importance (top 15) : MeanDecreaseAccuracy across all classes
imp_df <- as.data.frame(importance(model_rf, type = 1))
imp_df$Feature <- rownames(imp_df)
p_rf_varimp <- imp_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(Feature, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_col(fill = "#2E74B5") +
  coord_flip() +
  labs(title = "Random Forest : Variable Importance (Top 15)",
       x = NULL, y = "Mean Decrease in Accuracy") +
  theme_minimal(base_size = 12)
print(p_rf_varimp)
ggsave("plots/5_3_random_forest_variable_importance_top15.png", p_rf_varimp, width = 7, height = 5, dpi = 150)
```

#### 5.4 XGBoost

```{r model-xgb, echo=TRUE, results='hide', fig.show='hide'}
# OHE via model.matrix(); class weights via obs_weights; defaults: depth=6, eta=0.3, rounds=100

target_col <- "Degree of crash"

x_train_raw <- df_train_bal %>% select(-all_of(target_col))

x_test_raw <- df_test %>% select(-all_of(target_col))

y_train_factor <- df_train_bal[[target_col]]

y_test_factor  <- df_test[[target_col]]

y_train_factor <- factor(
  y_train_factor,
  levels = c("Fatal", "Injury", "Non_casualty")
)

y_test_factor <- factor(
  y_test_factor,
  levels = levels(y_train_factor)
)

class_levels <- levels(y_train_factor)

# XGBoost requires labels as 0, 1, 2

y_train <- as.integer(y_train_factor) - 1

y_test  <- as.integer(y_test_factor) - 1

table(y_train_factor)

train_crash_hour_median <- median(x_train_raw$crash_hour, na.rm = TRUE)

x_train_raw <- x_train_raw %>% mutate(
    crash_hour = ifelse(
      is.na(crash_hour),
      train_crash_hour_median,
      crash_hour
    )
  )

x_test_raw <- x_test_raw %>%
  mutate(
    crash_hour = ifelse(
      is.na(crash_hour),
      train_crash_hour_median,
      crash_hour
    )
  )

x_train_mm <- model.matrix(~ . - 1, data = x_train_raw)

x_test_mm  <- model.matrix(~ . - 1, data = x_test_raw)

# Clean feature names for XGBoost compatibility

colnames(x_train_mm) <- make.names(colnames(x_train_mm), unique = TRUE)

colnames(x_test_mm)  <- make.names(colnames(x_test_mm), unique = TRUE)

# Align test columns with training columns

missing_cols <- setdiff(colnames(x_train_mm), colnames(x_test_mm))

if (length(missing_cols) > 0) {
  missing_matrix <- matrix(
    0,
    nrow = nrow(x_test_mm),
    ncol = length(missing_cols)
  )
  colnames(missing_matrix) <- missing_cols
  x_test_mm <- cbind(x_test_mm, missing_matrix)

}

extra_cols <- setdiff(colnames(x_test_mm), colnames(x_train_mm))

if (length(extra_cols) > 0) {
  x_test_mm <- x_test_mm[, !(colnames(x_test_mm) %in% extra_cols), drop = FALSE]
}

x_test_mm <- x_test_mm[, colnames(x_train_mm), drop = FALSE]

# Final checks

stopifnot(nrow(x_train_mm) == length(y_train))

stopifnot(nrow(x_test_mm) == length(y_test))

stopifnot(identical(colnames(x_train_mm), colnames(x_test_mm)))

stopifnot(!any(is.na(x_train_mm)))

stopifnot(!any(is.na(x_test_mm)))

# XGBoost requires per-observation weights passed directly to xgb.DMatrix - it cannot consume caret's
# trainControl class_weights. Recalculating here from y_train_factor (derived from df_train_bal) gives
# the same inverse-frequency weighting as the preprocessing class_weights object, just scoped to this
# model's integer-encoded label vector. Intentionally overwrites the global class_weights name so that
# subsequent per-obs weight vectors (obs_weights_sub, obs_weights_full) use the correct XGBoost scale.
class_counts <- table(y_train_factor)

class_weights <- sum(class_counts) / (length(class_counts) * class_counts)

obs_weights <- as.numeric(class_weights[y_train_factor])

stopifnot(length(obs_weights) == nrow(x_train_mm))

stopifnot(!any(is.na(obs_weights)))

stopifnot(!any(is.infinite(obs_weights)))

dtrain <- xgb.DMatrix(
  data = x_train_mm,
  label = y_train,
  weight = obs_weights
)

dtest <- xgb.DMatrix(
  data = x_test_mm,
  label = y_test
)

xgb_params <- list(
  objective = "multi:softprob",
  num_class = length(class_levels),
  eval_metric = "mlogloss",
  max_depth = 6,
  eta = 0.3,
  gamma = 0,
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = 1
)

xgb_fit <- xgb.train(
  params = xgb_params,
  data = dtrain,
  nrounds = 100,
  verbose = 0
)

xgb_prob <- predict(xgb_fit, dtest)

xgb_prob_matrix <- matrix(
  xgb_prob,
  ncol = length(class_levels),
  byrow = TRUE
)

colnames(xgb_prob_matrix) <- class_levels

xgb_pred_index <- max.col(xgb_prob_matrix) - 1

xgb_pred <- factor(
  class_levels[xgb_pred_index + 1],
  levels = class_levels
)

cm_xgb <- confusionMatrix(
  data = xgb_pred,
  reference = y_test_factor
)

accuracy_xgb <- as.numeric(cm_xgb$overall["Accuracy"])

kappa_xgb    <- as.numeric(cm_xgb$overall["Kappa"])

by_class_xgb <- as.data.frame(cm_xgb$byClass) %>% rownames_to_column("Class")

macro_f1_xgb <- mean(
  by_class_xgb$F1,
  na.rm = TRUE
)

fatal_recall_xgb <- by_class_xgb %>% filter(Class == "Class: Fatal") %>% pull(Recall)

weighted_f1_xgb <- weighted.mean(
  by_class_xgb$F1,
  prop.table(table(y_test_factor))[gsub("Class: ", "", by_class_xgb$Class)],
  na.rm = TRUE
)

xgb_metrics <- tibble(
  model = "XGBoost default",
  accuracy = accuracy_xgb,
  macro_f1 = macro_f1_xgb,
  fatal_recall = fatal_recall_xgb,
  weighted_f1 = weighted_f1_xgb,
  kappa = kappa_xgb
)

p_xgb_cm <- as.data.frame(cm_xgb$table) %>% ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "XGBoost - Confusion Matrix",
    subtitle = paste0(
      "Fatal Recall: ", round(fatal_recall_xgb, 3),
      "  |  Macro-F1: ", round(macro_f1_xgb, 3)
    ),
    x = "Actual", y = "Predicted"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 10))

dir.create("plots", showWarnings = FALSE)

ggsave(
  "plots/5_4_xgboost_confusion_matrix.png",
  p_xgb_cm,
  width = 7,
  height = 5,
  dpi = 150
)

xgb_importance <- xgb.importance(
  feature_names = colnames(x_train_mm),
  model = xgb_fit
)

xgb_importance_top15 <- xgb_importance %>% as_tibble() %>% slice_head(n = 15)

p_xgb_varimp <- xgb_importance_top15 %>% ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "#2E74B5") +
  coord_flip() +
  labs(
    title = "XGBoost: Variable Importance (Top 15)",
    x = NULL,
    y = "Gain"
  ) +
  theme_minimal(base_size = 12)

ggsave(
  "plots/5_4_xgboost_variable_importance_top15.png",
  p_xgb_varimp,
  width = 7,
  height = 5,
  dpi = 150
)
```

#### 5.5 KNN

```{r model-knn, echo=TRUE, results='hide', fig.show='hide'}
# OHE via model.matrix(); k = round(sqrt(p)); Euclidean distance on OHE is a known limitation

knn_target      <- df_train_bal$`Degree of crash`
knn_target_test <- df_test$`Degree of crash`

df_combined_knn <- bind_rows(
  df_train_bal %>% mutate(.split = "train"),
  df_test      %>% mutate(.split = "test")
) %>% mutate(.row_id = row_number())

df_for_mm <- df_combined_knn %>%
  select(-`Degree of crash`, -.split, -.row_id) %>%
  mutate(across(where(is.factor), droplevels))

mm_combined <- model.matrix(~ . - 1, data = df_for_mm)

kept_ids    <- as.integer(rownames(mm_combined))
split_kept  <- df_combined_knn$.split[kept_ids]
target_kept <- df_combined_knn$`Degree of crash`[kept_ids]

knn_x_train <- mm_combined[split_kept == "train", ]
knn_x_test  <- mm_combined[split_kept == "test",  ]

knn_target      <- target_kept[split_kept == "train"]
knn_target_test <- target_kept[split_kept == "test"]

knn_train_df <- as.data.frame(knn_x_train); knn_train_df$outcome <- knn_target
knn_test_df  <- as.data.frame(knn_x_test);  knn_test_df$outcome  <- knn_target_test

# Set k = round(sqrt(p)) where p = number of features in the one-hot encoded matrix
p_knn <- ncol(knn_x_train)
k_knn <- round(sqrt(p_knn))
cat(sprintf("KNN: p = %d features  =>  k = round(sqrt(%d)) = %d\n", p_knn, p_knn, k_knn))

model_knn <- train(
  outcome ~ ., data = knn_train_df,
  method     = "kknn",
  preProcess = c("center", "scale"),
  trControl  = trainControl(method = "none"),   # direct fit - no CV, no SMOTE (df_train_bal already balanced)
  tuneGrid   = expand.grid(kmax = k_knn, distance = 2, kernel = "optimal")
)

pred_knn         <- predict(model_knn, newdata = knn_test_df)
cm_knn           <- confusionMatrix(pred_knn, knn_target_test)
fatal_recall_knn <- as.numeric(cm_knn$byClass["Class: Fatal", "Recall"])
f1_knn_vec       <- setNames(cm_knn$byClass[, "F1"], gsub("Class: ", "", rownames(cm_knn$byClass)))
test_props_knn   <- prop.table(table(knn_target_test))
macro_f1_knn     <- mean(f1_knn_vec, na.rm = TRUE)
weighted_f1_knn  <- weighted.mean(f1_knn_vec[names(test_props_knn)], test_props_knn, na.rm = TRUE)
kappa_knn        <- as.numeric(cm_knn$overall["Kappa"])

cat(sprintf(
  "KNN (k=%d, sqrt(p))  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  k_knn, macro_f1_knn, fatal_recall_knn, weighted_f1_knn, kappa_knn
))

p_knn_cm <- as.data.frame(cm_knn$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title    = paste0("KNN (k=", k_knn, ") - Confusion Matrix"),
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_knn, 3),
                      "  |  Macro-F1: ", round(macro_f1_knn, 3)),
    x = "Actual", y = "Predicted"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 10))
print(p_knn_cm)
ggsave("plots/5_5_knn_confusion_matrix.png", p_knn_cm, width = 7, height = 5, dpi = 150)
```

#### 5.6 Naive Bayes

```{r model-nb, echo=TRUE, results='hide', fig.show='hide'}
# SMOTE inside CV folds; usekernel=FALSE (kernel density unreliable on ~770 Fatal rows)

# Safety check: create df_train_bal if it does not exist
if (!exists("df_train_bal")) {
  fatal_rows   <- df_train %>% filter(`Degree of crash` == "Fatal")
  injury_rows  <- df_train %>% filter(`Degree of crash` == "Injury") %>% slice_sample(n = 15000)
  noncas_rows  <- df_train %>% filter(`Degree of crash` == "Non_casualty") %>% slice_sample(n = 15000)
  df_train_bal <- bind_rows(fatal_rows, injury_rows, noncas_rows) %>% slice_sample(prop = 1)
}

# Prepare data for Naive Bayes
df_train_nb <- df_train_bal
df_test_nb  <- df_test

# Remove columns that are entirely NA
df_train_nb <- df_train_nb[, colSums(is.na(df_train_nb)) < nrow(df_train_nb)]
df_test_nb  <- df_test_nb[, names(df_train_nb)]

# Remove rows with missing target
df_train_nb <- df_train_nb %>% filter(!is.na(`Degree of crash`))
df_test_nb  <- df_test_nb %>% filter(!is.na(`Degree of crash`))

# Convert character predictors to factors
df_train_nb <- df_train_nb %>%
  mutate(across(where(is.character), as.factor))

df_test_nb <- df_test_nb %>%
  mutate(across(where(is.character), as.factor))

# Convert predictor NA to "Unknown", but do not touch the target
predictor_cols <- setdiff(names(df_train_nb), "Degree of crash")

for (col in predictor_cols) {
  if (is.factor(df_train_nb[[col]])) {
    df_train_nb[[col]] <- fct_na_value_to_level(df_train_nb[[col]], level = "Unknown")
    df_test_nb[[col]]  <- fct_na_value_to_level(df_test_nb[[col]], level = "Unknown")
  }
}

# Remove remaining rows with numeric NA
df_train_nb <- na.omit(df_train_nb)
df_test_nb  <- na.omit(df_test_nb)

# Drop unused target levels
df_train_nb$`Degree of crash` <- droplevels(as.factor(df_train_nb$`Degree of crash`))
df_test_nb$`Degree of crash`  <- droplevels(as.factor(df_test_nb$`Degree of crash`))

# Train baseline Naive Bayes
nb_model <- train(
  `Degree of crash` ~ .,
  data = df_train_nb,
  method = "naive_bayes",
  trControl = cv_ctrl_smote,
  metric = "logLoss",
  tuneGrid = expand.grid(
    laplace = 0,
    usekernel = FALSE,
    adjust = 1
  )
)

nb_model$results

# Prediction
pred_nb <- predict(nb_model, df_test_nb)

# Confusion matrix
cm_nb <- confusionMatrix(
  pred_nb,
  df_test_nb$`Degree of crash`
)

cm_nb

# Naive Bayes Performance Metrics


recall_nb <- cm_nb$byClass[, "Sensitivity"]
precision_nb <- cm_nb$byClass[, "Pos Pred Value"]

f1_nb <- 2 * precision_nb * recall_nb / (precision_nb + recall_nb)
f1_nb[is.na(f1_nb)] <- 0

macro_f1_nb <- mean(f1_nb, na.rm = TRUE)
fatal_recall_nb <- recall_nb["Class: Fatal"]

class_weights <- prop.table(table(df_test_nb$`Degree of crash`))
f1_names_clean <- gsub("Class: ", "", names(f1_nb))

weighted_f1_nb <- sum(
  f1_nb * class_weights[f1_names_clean],
  na.rm = TRUE
)

kappa_nb <- cm_nb$overall["Kappa"]

nb_metrics <- data.frame(
  Model = "Naive Bayes",
  Macro_F1 = round(macro_f1_nb, 3),
  Fatal_Recall = round(fatal_recall_nb, 3),
  Weighted_F1 = round(weighted_f1_nb, 3),
  Kappa = round(kappa_nb, 3)
)

knitr::kable(
  nb_metrics,
  caption = "Naive Bayes Performance Metrics"
)

# Naive Bayes Confusion Matrix Heatmap

cm_nb_df <- as.data.frame(cm_nb$table)

p_nb_cm <- ggplot(cm_nb_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Naive Bayes - Confusion Matrix",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_nb, 3),
                      "  |  Macro-F1: ", round(macro_f1_nb, 3)),
    x = "Actual", y = "Predicted",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 10))

p_nb_cm
ggsave("plots/5_6_nb_confusion_matrix.png", p_nb_cm, width = 7, height = 5, dpi = 150)
```

```{r results-df, echo=TRUE, results='hide'}
# 5.7 Results summary dataframe
results_df <- data.frame(
  Model        = c("Logistic Regression", "Decision Tree", "Random Forest",
                   "XGBoost", paste0("KNN (k=", k_knn, ")"), "Naive Bayes"),
  Macro_F1     = round(c(macro_f1_lr,  macro_f1_dt,  macro_f1_rf,
                         macro_f1_xgb, macro_f1_knn, macro_f1_nb), 4),
  Fatal_Recall = round(c(fatal_recall_lr,  fatal_recall_dt,  as.numeric(fatal_recall_rf),
                         fatal_recall_xgb, fatal_recall_knn, as.numeric(fatal_recall_nb)), 4),
  Weighted_F1  = round(c(weighted_f1_lr,  weighted_f1_dt,  as.numeric(weighted_f1_rf),
                         weighted_f1_xgb, weighted_f1_knn, weighted_f1_nb), 4),
  Kappa        = round(c(kappa_lr,  kappa_dt,  as.numeric(kappa_rf),
                         kappa_xgb, kappa_knn, as.numeric(kappa_nb)), 4)
)
```

```{r results-table, echo=TRUE, results='asis'}
results_display <- results_df %>%
  mutate(
    Fatal_Recall_Constraint = ifelse(Fatal_Recall >= 0.50, "PASS", "FAIL")
  )

knitr::kable(
  results_display,
  col.names = c("Model", "Macro-F1", "Fatal Recall", "Weighted F1", "Kappa", "Fatal Recall ≥ 0.50"),
  align = c("l", "r", "r", "r", "r", "c"),
  caption = "Table 5: Baseline Model Performance on the Test Set"
) |>
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "condensed"),
    font_size = 12
  ) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::row_spec(3, bold = TRUE, background = "#FFF3CD") |>
  kableExtra::row_spec(5, bold = TRUE, background = "#D6EAF8") |>
  kableExtra::column_spec(1, bold = TRUE) |>
  kableExtra::column_spec(3, color = "#C0392B")
```

**Safety constraint (Fatal Recall ≥ 0.50):**

- **Fail - XGBoost** (Fatal Recall = `r round(fatal_recall_xgb, 3)`): collapses to near-random discrimination (Kappa = `r round(kappa_xgb, 4)`), with the one-hot expansion of high-cardinality categoricals interacting poorly with the inverse-frequency weighting scheme
- **Fail - Random Forest** (Fatal Recall = `r round(as.numeric(fatal_recall_rf), 3)`): narrowly misses the floor at baseline despite leading all models on Macro-F1 (`r round(macro_f1_rf, 3)`) and Kappa (`r round(as.numeric(kappa_rf), 3)`), addressed by mtry tuning in Section 6

**Models that pass the safety floor:**

- **LR** (Fatal Recall = `r round(fatal_recall_lr, 3)`) and **DT** (Fatal Recall = `r round(fatal_recall_dt, 3)`): high recall at the cost of Macro-F1 (`r round(macro_f1_lr, 3)` and `r round(macro_f1_dt, 3)` respectively) - both over-predict Fatal, producing many false alarms in Injury and Non-casualty
- **NB** (Fatal Recall = `r round(as.numeric(fatal_recall_nb), 3)`): passes the floor but achieves the lowest Macro-F1 (`r round(macro_f1_nb, 3)`) and Weighted F1 (`r round(weighted_f1_nb, 3)`) of all models, as the conditional independence assumption is badly violated in this dataset
- **KNN** (Fatal Recall = `r round(fatal_recall_knn, 3)`, Macro-F1 = `r round(macro_f1_knn, 3)`): the only baseline model to satisfy **both** constraints simultaneously, distributing predictions across all three classes more evenly than the parametric models

```{r cm-grid, echo=TRUE, results='show', fig.width=14, fig.height=9}
plot_grid(p_lr_cm, p_dt_cm, p_rf_cm,
          p_xgb_cm, p_knn_cm, p_nb_cm,
          ncol = 3, nrow = 2,
          labels = c("LR", "DT", "RF", "XGBoost", "KNN", "NB"),
          label_size = 11)
```

**Confusion matrix patterns:**

- **RF & KNN** distribute predictions across all three classes with the diagonal cell dominant in every row, indicating genuine multi-class discrimination rather than collapsing to a majority-class default
- **LR, DT & NB** achieve high Fatal Recall by aggressively predicting Fatal, with the Fatal predicted row containing a large share of misclassified Injury and Non-casualty instances (poor Fatal precision), which suppresses per-class F1 and drags down Macro-F1
- **XGBoost** spreads predictions roughly equally across all three classes but with very low accuracy in each row, producing near-random discrimination (Kappa ≈ `r round(kappa_xgb, 3)`), suggesting the one-hot expansion of high-cardinality categoricals interacts poorly with the inverse-frequency weighting scheme
- **NB** achieves high Fatal Recall but its Weighted F1 (`r round(weighted_f1_nb, 3)`) is the lowest of all models, confirming that its high recall comes at the cost of severely misclassifying Injury and Non-casualty instances

------------------------------------------------------------------------

### 6 Model Improvement

Each model's primary hyperparameter was tuned using the same 5-fold CV framework from Section 4: L2 weight decay for LR, CP pruning for DT, `tuneRF()` for RF, a grid over `nrounds`/`max_depth`/`eta` for XGBoost, and Laplace smoothing for NB. KNN required no additional tuning ($k = \lfloor\sqrt{p}\rfloor$ set by rule). All code is folded below, with a unified before/after comparison at the end of the section.

#### 6.1 Logistic Regression - L2 decay tuning

```{r improve-lr, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# Multinomial Logistic Regression  -  Improvement
# 1. Define the tuning grid
lr_grid <- expand.grid(decay = c(0, 0.001, 0.01, 0.1))

# 2. Model Training and Tuning (Based on Section 5.1)
model_lr_tuned <- train(
  Degree.of.crash ~ .,
  data = train_lr,
  method = "multinom",
  trControl = cv_ctrl_smote,
  tuneGrid = lr_grid,
  preProcess = "nzv",
  metric = "Mean_F1",
  trace = FALSE,
  MaxNWts = 20000
)

best_decay <- model_lr_tuned$bestTune$decay
cat("Best decay parameter from tuning:", best_decay, "\n")

# 3. Plot the decay vs Mean_F1 curve
# transform decay into factor to ensure the uniform distribution of the horizontal axis
p_lr_tune <- ggplot(model_lr_tuned$results, aes(x = as.factor(decay), y = Mean_F1, group = 1)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 3, colour = "#2E74B5") +
  geom_vline(xintercept = which(lr_grid$decay == best_decay), linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = which(lr_grid$decay == best_decay), y = max(model_lr_tuned$results$Mean_F1),
           label = paste("Best decay =", best_decay),
           hjust = -0.2, colour = "#C0392B", size = 3.5) +
  labs(
    title = "LR Tuning: Weight Decay vs Macro-F1",
    x = "Weight Decay (L2 Regularization Penalty)",
    y = "Cross-Validated Macro-F1"
  ) +
  theme_minimal(base_size = 12)

print(p_lr_tune)
ggsave("plots/6_1_logistic_regression_decay_tuning_curve.png", p_lr_tune, width = 7, height = 5, dpi = 150)

# 4. Make predictions on the test set and calculate new metrics
pred_lr_tuned <- predict(model_lr_tuned, newdata = test_lr)
cm_lr_tuned <- confusionMatrix(pred_lr_tuned, test_lr$Degree.of.crash)

fatal_recall_lr_tuned <- as.numeric(cm_lr_tuned$byClass["Class: Fatal", "Recall"])
f1_lr_tuned_vec <- setNames(cm_lr_tuned$byClass[, "F1"], gsub("Class: ", "", rownames(cm_lr_tuned$byClass)))
test_props_lr_tuned <- prop.table(table(test_lr$Degree.of.crash))
macro_f1_lr_tuned   <- mean(f1_lr_tuned_vec, na.rm = TRUE)
weighted_f1_lr_tuned <- weighted.mean(f1_lr_tuned_vec[names(test_props_lr_tuned)],
                                      test_props_lr_tuned, na.rm = TRUE)
kappa_lr_tuned <- as.numeric(cm_lr_tuned$overall["Kappa"])

cat(sprintf(
  "   Logistic Regression (Tuned)  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_lr_tuned, fatal_recall_lr_tuned, weighted_f1_lr_tuned, kappa_lr_tuned
))

# 5. Before / After Comparison Table
before_after_lr <- data.frame(
  Model = c("LR: default decay (0)", paste0("LR: tuned decay (", best_decay, ")")),
  `Macro-F1` = round(c(macro_f1_lr, macro_f1_lr_tuned), 4),
  `Fatal Recall` = round(c(fatal_recall_lr, fatal_recall_lr_tuned), 4),
  check.names = FALSE
)
knitr::kable(before_after_lr, caption = "Logistic Regression  -  Before vs After Tuning")

p_lr_tuned_cm <- as.data.frame(cm_lr_tuned$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Logistic Regression (Tuned) - Confusion Matrix",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_lr_tuned, 3),
                      "  |  Macro-F1: ", round(macro_f1_lr_tuned, 3)),
    x = "Actual", y = "Predicted"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 10))
ggsave("plots/6_1_logistic_regression_tuned_confusion_matrix.png", p_lr_tuned_cm, width = 7, height = 5, dpi = 150)
```

#### 6.2 Decision Tree - CP pruning

```{r improve-dt, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# Decision Tree improvement  -  tune complexity by pruning via CP table
# 1. First grow a deeper tree (relax the cp parameter to 0.001) to find the optimal split points
model_dt_deep <- rpart(
  Degree.of.crash ~ .,
  data = train_dt,
  method = "class",
  weights = dt_weights,
  control = rpart.control(cp = 0.001, xval = 5) # xval=5
)

# 2. Extract the CP with minimum xerror from the deep tree
best_cp <- model_dt_deep$cptable[which.min(model_dt_deep$cptable[, "xerror"]), "CP"]
cat("Best cp from xerror minimisation:", best_cp, "\n")

# 3. ggplot CP tuning curve (used in tuning-curves grid - avoids draw_image/magick dependency)
best_nsplit <- model_dt_deep$cptable[which.min(model_dt_deep$cptable[, "xerror"]), "nsplit"]
p_dt_tune <- as.data.frame(model_dt_deep$cptable) %>%
  ggplot(aes(x = nsplit, y = xerror)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 2.5, colour = "#2E74B5") +
  geom_errorbar(aes(ymin = xerror - xstd, ymax = xerror + xstd), width = 0.3, alpha = 0.4) +
  geom_vline(xintercept = best_nsplit, linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = best_nsplit, y = max(model_dt_deep$cptable[, "xerror"]),
           label = paste("Best cp =", signif(best_cp, 3)),
           hjust = -0.1, colour = "#C0392B", size = 3.5) +
  labs(title = "DT Tuning: Splits vs CV Error",
       x = "Number of Splits", y = "CV Error (xerror)") +
  theme_minimal(base_size = 12)
ggsave("plots/6_2_decision_tree_cp_plot.png", p_dt_tune, width = 7, height = 5, dpi = 150)

# 4. Prune the deep tree using the optimal CP found
model_dt_tuned <- prune(model_dt_deep, cp = best_cp)

# 5. Evaluate the pruned model on the test set
pred_dt_tuned <- predict(model_dt_tuned, newdata = test_dt, type = "class")
cm_dt_tuned <- confusionMatrix(pred_dt_tuned, test_dt$Degree.of.crash)

fatal_recall_dt_tuned <- as.numeric(cm_dt_tuned$byClass["Class: Fatal", "Recall"])
f1_dt_tuned_vec <- setNames(cm_dt_tuned$byClass[, "F1"], gsub("Class: ", "", rownames(cm_dt_tuned$byClass)))
macro_f1_dt_tuned <- mean(f1_dt_tuned_vec, na.rm = TRUE)
weighted_f1_dt_tuned <- weighted.mean(f1_dt_tuned_vec[names(test_props_dt)], test_props_dt, na.rm = TRUE)
kappa_dt_tuned <- as.numeric(cm_dt_tuned$overall["Kappa"])

cat(sprintf(
  "   Decision Tree (Tuned)  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_dt_tuned, fatal_recall_dt_tuned, weighted_f1_dt_tuned, kappa_dt_tuned
))

# 6. Generate Before/After comparison table
before_after_dt <- data.frame(
  Model = c("DT: default cp (0.01)", paste0("DT: tuned cp (", signif(best_cp, 3), ")")),
  `Macro-F1` = round(c(macro_f1_dt, macro_f1_dt_tuned), 4),
  `Fatal Recall` = round(c(fatal_recall_dt, fatal_recall_dt_tuned), 4),
  check.names = FALSE
)
knitr::kable(before_after_dt, caption = "Decision Tree  -  Before vs After Pruning")

# 7. Plot Tuned Confusion Matrix heatmap
p_dt_tuned_cm <- as.data.frame(cm_dt_tuned$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Decision Tree (Tuned) - Confusion Matrix",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_dt_tuned, 3),
                      "  |  Macro-F1: ", round(macro_f1_dt_tuned, 3)),
    x = "Actual", y = "Predicted"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 10))
print(p_dt_tuned_cm)
ggsave("plots/6_2_decision_tree_tuned_confusion_matrix.png", p_dt_tuned_cm, width = 7, height = 5, dpi = 150)
```

#### 6.3 Random Forest - mtry tuning

```{r improve-rf, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
invisible(capture.output(
  tune_result <- tuneRF(
    x          = rf_x_train,
    y          = df_train_bal$`Degree of crash`,
    mtryStart  = floor(sqrt(ncol(rf_x_train))),
    ntreeTry   = 50,
    stepFactor = 1.5,
    improve    = 0.01,
    trace      = FALSE,
    plot       = FALSE
  )
))
best_mtry <- tune_result[which.min(tune_result[, "OOBError"]), "mtry"]
cat("Best mtry from tuneRF:", best_mtry, "\n")

# mtry vs OOB error curve
p_rf_tune <- as.data.frame(tune_result) %>%
  ggplot(aes(x = mtry, y = OOBError)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 3, colour = "#2E74B5") +
  geom_vline(xintercept = best_mtry, linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = best_mtry, y = max(tune_result[, "OOBError"]),
           label = paste("Best mtry =", best_mtry),
           hjust = 1.1, colour = "#C0392B", size = 3.5) +
  labs(title = "RF Tuning: mtry vs OOB Error",
       x = "mtry (variables randomly sampled per split)", y = "OOB Error Rate") +
  theme_minimal(base_size = 12)
print(p_rf_tune)
ggsave("plots/6_3_random_forest_mtry_tuning_curve.png", p_rf_tune, width = 7, height = 5, dpi = 150)

# Final tuned RF - df_train_bal (balanced) with best_mtry from tuneRF
model_rf_tuned <- randomForest(
  x          = rf_x_train,
  y          = df_train_bal$`Degree of crash`,
  mtry       = best_mtry,
  ntree      = 100,
  importance = TRUE
)

pred_rf_tuned      <- predict(model_rf_tuned, rf_x_test)
cm_rf_tuned        <- confusionMatrix(pred_rf_tuned, df_test$`Degree of crash`)
fatal_recall_tuned <- cm_rf_tuned$byClass["Class: Fatal", "Recall"]
macro_f1_tuned     <- mean(cm_rf_tuned$byClass[, "F1"], na.rm = TRUE)
weighted_f1_tuned  <- weighted.mean(cm_rf_tuned$byClass[, "F1"],
                       prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_tuned        <- cm_rf_tuned$overall["Kappa"]

cat(sprintf(
  "    RF Tuned   -   Test Set Metrics \nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf_tuned$err.rate[100, "OOB"], macro_f1_tuned, fatal_recall_tuned, weighted_f1_tuned, kappa_tuned
))

# Before / After comparison table
data.frame(
  Model          = c("RF : default mtry", "RF : tuned mtry"),
  mtry           = c(floor(sqrt(ncol(rf_x_train))), best_mtry),
  `Macro-F1`     = round(c(macro_f1_rf,    macro_f1_tuned),    4),
  `Fatal Recall` = round(c(fatal_recall_rf, fatal_recall_tuned), 4),
  check.names = FALSE
) %>% knitr::kable(caption = "Random Forest  -  Before vs After Tuning")

# Plot 2: OOB error vs ntree convergence curve
p_rf_oob <- data.frame(ntree = seq_len(nrow(model_rf_tuned$err.rate)),
           OOB   = model_rf_tuned$err.rate[, "OOB"]) %>%
  ggplot(aes(x = ntree, y = OOB)) +
  geom_line(colour = "#2E74B5") +
  labs(title = "Random Forest : OOB Error vs Number of Trees (tuned mtry)",
       x = "Number of trees", y = "OOB error rate") +
  theme_minimal(base_size = 12)
print(p_rf_oob)
ggsave("plots/6_3_random_forest_oob_convergence.png", p_rf_oob, width = 7, height = 5, dpi = 150)

# Plot 3: Confusion matrix heatmap  -  tuned model (last in row)
p_rf_tuned_cm <- as.data.frame(cm_rf_tuned$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(title = "Random Forest (Tuned) - Confusion Matrix",
       subtitle = paste0("Fatal Recall: ", round(fatal_recall_tuned, 3),
                         "  |  Macro-F1: ", round(macro_f1_tuned, 3)),
       x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 10))
print(p_rf_tuned_cm)
ggsave("plots/6_3_random_forest_tuned_confusion_matrix.png", p_rf_tuned_cm, width = 7, height = 5, dpi = 150)
```

#### 6.4 XGBoost - grid search with early stopping

```{r improve-xgb, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
valid_idx <- createDataPartition(
  y_train_factor,
  p = 0.2,
  list = FALSE
)

x_valid_mm <- x_train_mm[valid_idx, , drop = FALSE]

y_valid_factor <- y_train_factor[valid_idx]

y_valid <- y_train[valid_idx]

x_subtrain_mm <- x_train_mm[-valid_idx, , drop = FALSE]

y_subtrain_factor <- y_train_factor[-valid_idx]

y_subtrain <- y_train[-valid_idx]

class_counts_sub <- table(y_subtrain_factor)

class_weights_sub <- sum(class_counts_sub) / (length(class_counts_sub) * class_counts_sub)

obs_weights_sub <- as.numeric(class_weights_sub[y_subtrain_factor])

dsubtrain <- xgb.DMatrix(
  data = x_subtrain_mm,
  label = y_subtrain,
  weight = obs_weights_sub
)

dvalid <- xgb.DMatrix(
  data = x_valid_mm,
  label = y_valid
)

# Small grid: 2 depths x 3 etas = 6 combinations
# nrounds is determined automatically via early stopping (patience = 20 rounds)
xgb_tune_grid <- expand.grid(
  max_depth = c(3, 5),
  eta       = c(0.05, 0.1, 0.3)
)

evaluate_xgb_grid <- function(max_depth, eta) {
  params <- list(
    objective        = "multi:softprob",
    num_class        = length(class_levels),
    eval_metric      = "mlogloss",
    max_depth        = max_depth,
    eta              = eta,
    gamma            = 0,
    colsample_bytree = 1,
    min_child_weight = 1,
    subsample        = 1
  )

  model <- xgb.train(
    params              = params,
    data                = dsubtrain,
    nrounds             = 500,
    watchlist           = list(val = dvalid),
    early_stopping_rounds = 20,
    verbose             = 0
  )

  bi <- model$best_iteration
  best_nrounds <- if (is.null(bi) || length(bi) == 0 || is.na(bi) || bi < 1L) 500L else as.integer(bi)

  pred_prob        <- predict(model, dvalid)
  pred_prob_matrix <- matrix(pred_prob, ncol = length(class_levels), byrow = TRUE)
  pred_index       <- max.col(pred_prob_matrix) - 1
  pred_class       <- factor(class_levels[pred_index + 1], levels = class_levels)

  cm       <- confusionMatrix(data = pred_class, reference = y_valid_factor)
  by_class <- as.data.frame(cm$byClass)
  by_class$Class <- rownames(by_class)
  rownames(by_class) <- NULL

  tibble(
    max_depth    = max_depth,
    eta          = eta,
    best_nrounds = best_nrounds,
    Accuracy     = as.numeric(cm$overall["Accuracy"]),
    Kappa        = as.numeric(cm$overall["Kappa"]),
    Mean_F1      = mean(by_class$F1, na.rm = TRUE),
    Fatal_Recall = by_class %>% filter(Class == "Class: Fatal") %>% pull(Sensitivity)
  )
}

xgb_tune_results <- bind_rows(
  lapply(seq_len(nrow(xgb_tune_grid)), function(i) {
    evaluate_xgb_grid(
      max_depth = xgb_tune_grid$max_depth[i],
      eta       = xgb_tune_grid$eta[i]
    )
  })
)

best_xgb_params <- xgb_tune_results %>%
  arrange(desc(Mean_F1), desc(Fatal_Recall)) %>%
  slice(1)

# Tuning plot: Mean-F1 for each setting (colour = eta, x = max_depth)
p_xgb_tune_curve <- xgb_tune_results %>%
  mutate(setting = paste0("depth=", max_depth, ", eta=", eta),
         eta_lbl = factor(paste0("eta=", eta))) %>%
  ggplot(aes(x = factor(max_depth), y = Mean_F1, fill = eta_lbl)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(Mean_F1, 3)), position = position_dodge(0.9),
            vjust = 1.4, size = 2.6, colour = "white", fontface = "bold") +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  labs(title = "XGBoost Tuning: Mean-F1 by Depth & Learning Rate",
       x = "max_depth", y = "Validation Mean-F1", fill = "Learning rate") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(size = 10))

dir.create("plots", showWarnings = FALSE)

ggsave(
  "plots/6_4_xgboost_tuning_curve.png",
  p_xgb_tune_curve,
  width = 7,
  height = 5,
  dpi = 150
)

class_counts_full <- table(y_train_factor)

class_weights_full <- sum(class_counts_full) /
  (length(class_counts_full) * class_counts_full)

obs_weights_full <- as.numeric(class_weights_full[y_train_factor])

dtrain_full <- xgb.DMatrix(
  data = x_train_mm,
  label = y_train,
  weight = obs_weights_full
)

dtest_full <- xgb.DMatrix(
  data = x_test_mm,
  label = y_test
)

xgb_tuned_params <- list(
  objective = "multi:softprob",
  num_class = length(class_levels),
  eval_metric = "mlogloss",
  max_depth = best_xgb_params$max_depth,
  eta = best_xgb_params$eta,
  gamma = 0,
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = 1
)

xgb_fit_tuned <- xgb.train(
  params = xgb_tuned_params,
  data = dtrain_full,
  nrounds = max(1L, as.integer(best_xgb_params$best_nrounds)),
  verbose = 0
)

xgb_tuned_prob <- predict(xgb_fit_tuned, dtest_full)

xgb_tuned_prob_matrix <- matrix(
  xgb_tuned_prob,
  ncol = length(class_levels),
  byrow = TRUE
)

xgb_tuned_pred_index <- max.col(xgb_tuned_prob_matrix) - 1

xgb_tuned_pred <- factor(
  class_levels[xgb_tuned_pred_index + 1],
  levels = class_levels
)

cm_xgb_tuned <- confusionMatrix(
  data = xgb_tuned_pred,
  reference = y_test_factor
)

by_class_xgb_tuned <- as.data.frame(cm_xgb_tuned$byClass)

by_class_xgb_tuned$Class <- rownames(by_class_xgb_tuned)

rownames(by_class_xgb_tuned) <- NULL

accuracy_xgb_tuned <- as.numeric(cm_xgb_tuned$overall["Accuracy"])

kappa_xgb_tuned <- as.numeric(cm_xgb_tuned$overall["Kappa"])

macro_f1_xgb_tuned <- mean(
  by_class_xgb_tuned$F1,
  na.rm = TRUE
)

fatal_recall_xgb_tuned <- by_class_xgb_tuned %>%
  filter(Class == "Class: Fatal") %>%
  pull(Sensitivity)

weighted_f1_xgb_tuned <- weighted.mean(
  by_class_xgb_tuned$F1,
  prop.table(table(y_test_factor))[gsub("Class: ", "", by_class_xgb_tuned$Class)],
  na.rm = TRUE
)

xgb_before_after <- tibble(
  Model = c("XGBoost baseline", "XGBoost tuned"),
  Accuracy = c(accuracy_xgb, accuracy_xgb_tuned),
  Macro_F1 = c(macro_f1_xgb, macro_f1_xgb_tuned),
  Fatal_Recall = c(fatal_recall_xgb, fatal_recall_xgb_tuned),
  Weighted_F1 = c(weighted_f1_xgb, weighted_f1_xgb_tuned),
  Kappa = c(kappa_xgb, kappa_xgb_tuned)
)

xgb_before_after %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

p_xgb_tuned_cm <- as.data.frame(cm_xgb_tuned$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "XGBoost (Tuned) - Confusion Matrix",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_xgb_tuned, 3),
                      "  |  Macro-F1: ", round(macro_f1_xgb_tuned, 3)),
    x = "Actual", y = "Predicted"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(size = 10))
ggsave("plots/6_4_xgboost_tuned_confusion_matrix.png", p_xgb_tuned_cm, width = 7, height = 5, dpi = 150)
```

#### 6.5 KNN - No Additional Tuning

KNN was not subjected to grid search or cross-validated tuning. The primary reason is computational: repeated distance computations over a ~31,500 $\times$ `r p_knn` one-hot encoded matrix scale as $O(n \cdot p)$ per query point, making 5-fold CV over even a modest $k$-grid prohibitively slow within a single knit session.

Instead, $k = \lfloor\sqrt{p}\rfloor = `r k_knn`$ was set directly by a well-established rule of thumb. As $n$ grows large and $k/n$ shrinks to zero, the KNN error rate converges to twice the Bayes error rate (Cover & Hart, 1967). Choosing $k \propto \sqrt{p}$ balances the bias-variance trade-off across the feature dimensionality: a small $k$ overfits to local noise in the high-dimensional OHE space, while a large $k$ over-smooths class boundaries and loses sensitivity to the rare Fatal class. Hastie, Tibshirani & Friedman (2009, Section 13.3) endorse this heuristic as a practical starting point when the training set size and dimensionality are of comparable order, which is the case here.

Accordingly, the baseline metrics from Section 5 (Macro-F1 = `r round(macro_f1_knn, 3)`, Fatal Recall = `r round(fatal_recall_knn, 3)`) are carried forward as the final tuned KNN results.

#### 6.6 Naive Bayes - Laplace smoothing

```{r improve-nb, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# Naive Bayes Improvement

# Tuning control for Naive Bayes - use multiClassSummary + Mean_F1 (same as other models).
# mnLogLoss was previously used but rewards majority-class confidence, driving Fatal Recall to 0.
# usekernel = TRUE excluded: kernel density estimation on ~770 Fatal training rows
# produces unreliable density estimates that collapse Fatal predicted probabilities at test time.
ctrl_nb_tune <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = multiClassSummary,
  sampling = "smote",
  allowParallel = FALSE
)

# Small tuning grid - usekernel fixed to FALSE to avoid Fatal Recall collapse
nb_grid <- expand.grid(
  laplace    = c(0, 0.5, 1),
  usekernel  = FALSE,
  adjust     = c(1)
)

nb_tuned <- train(
  `Degree of crash` ~ .,
  data = df_train_nb,
  method = "naive_bayes",
  trControl = ctrl_nb_tune,
  metric = "Mean_F1",
  tuneGrid = nb_grid
)

# Best tuning parameters and CV results
nb_tuned$bestTune
nb_tuned$results

# Tuned Naive Bayes Evaluation


pred_nb_tuned <- predict(nb_tuned, df_test_nb)

cm_nb_tuned <- confusionMatrix(
  pred_nb_tuned,
  df_test_nb$`Degree of crash`
)

# Extract metrics
recall_nb_tuned <- cm_nb_tuned$byClass[, "Sensitivity"]
precision_nb_tuned <- cm_nb_tuned$byClass[, "Pos Pred Value"]

f1_nb_tuned <- 2 * precision_nb_tuned * recall_nb_tuned /
  (precision_nb_tuned + recall_nb_tuned)

f1_nb_tuned[is.na(f1_nb_tuned)] <- 0

macro_f1_nb_tuned <- mean(f1_nb_tuned, na.rm = TRUE)
fatal_recall_nb_tuned <- recall_nb_tuned["Class: Fatal"]

f1_names_tuned_clean <- gsub("Class: ", "", names(f1_nb_tuned))

weighted_f1_nb_tuned <- sum(
  f1_nb_tuned * class_weights[f1_names_tuned_clean],
  na.rm = TRUE
)

kappa_nb_tuned <- cm_nb_tuned$overall["Kappa"]

nb_tuned_metrics <- data.frame(
  Model = "Tuned Naive Bayes",
  Macro_F1 = round(macro_f1_nb_tuned, 3),
  Fatal_Recall = round(fatal_recall_nb_tuned, 3),
  Weighted_F1 = round(weighted_f1_nb_tuned, 3),
  Kappa = round(kappa_nb_tuned, 3)
)

knitr::kable(
  nb_tuned_metrics,
  caption = "Tuned Naive Bayes Performance Metrics"
)

# Before vs After Comparison


nb_before_after <- data.frame(
  Model = c("Baseline NB", "Tuned NB"),
  Macro_F1 = c(round(macro_f1_nb, 3), round(macro_f1_nb_tuned, 3)),
  Fatal_Recall = c(round(fatal_recall_nb, 3), round(fatal_recall_nb_tuned, 3)),
  Weighted_F1 = c(round(weighted_f1_nb, 3), round(weighted_f1_nb_tuned, 3)),
  Kappa = c(round(kappa_nb, 3), round(kappa_nb_tuned, 3))
)

knitr::kable(
  nb_before_after,
  caption = "Naive Bayes Before and After Hyperparameter Tuning"
)

# Compact Visualisation


# Tuning plot
p_nb_tune <- ggplot(
  nb_tuned$results,
  aes(
    x = factor(laplace),
    y = logLoss,
    fill = as.factor(usekernel)
  )
) +
  geom_col(position = "dodge", width = 0.7) +
  labs(
    title = "NB Tuning Results",
    x = "Laplace",
    y = "CV Log Loss",
    fill = "Kernel"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom"
  )

# Tuned confusion matrix heatmap
cm_nb_tuned_df <- as.data.frame(cm_nb_tuned$table)

p_nb_tuned_cm <- ggplot(
  cm_nb_tuned_df,
  aes(x = Reference, y = Prediction, fill = Freq)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Naive Bayes (Tuned) - Confusion Matrix",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_nb_tuned, 3),
                      "  |  Macro-F1: ", round(macro_f1_nb_tuned, 3)),
    x = "Actual",
    y = "Predicted",
    fill = "Count"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 9),
    legend.position = "right"
  )

ggsave("plots/6_6_nb_tuning_results.png", p_nb_tune, width = 5.5, height = 4, dpi = 220)
ggsave("plots/6_6_tuned_nb_confusion_matrix.png", p_nb_tuned_cm, width = 6, height = 4, dpi = 220)
```

```{r nb-tuning-display, echo=FALSE, results='hide', fig.show='hide', fig.width=11, fig.height=4}
plot_grid(p_nb_tune, p_nb_tuned_cm, ncol = 2)
```

```{r tuning-table, echo=TRUE, results='asis'}
tuning_compare <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest",
            "XGBoost", "KNN", "Naive Bayes"),
  Method = c(
    paste0("L2 decay; best = ", best_decay),
    paste0("CP pruning; best = ", signif(best_cp, 3)),
    paste0("tuneRF; best mtry = ", best_mtry),
    paste0("Grid; best nrounds/depth/eta = ",
           best_xgb_params$best_nrounds, "/",
           best_xgb_params$max_depth, "/",
           best_xgb_params$eta),
    paste0("k = round(sqrt(p)) = ", k_knn, " (rule)"),
    "Laplace; best = 0"
  ),
  Base_MF1  = round(c(macro_f1_lr,  macro_f1_dt,  macro_f1_rf,
                       macro_f1_xgb, macro_f1_knn, macro_f1_nb), 4),
  Tuned_MF1 = round(c(macro_f1_lr_tuned, macro_f1_dt_tuned,
                        as.numeric(macro_f1_tuned),
                        macro_f1_xgb_tuned, macro_f1_knn,
                        macro_f1_nb_tuned), 4),
  Base_FR   = round(c(fatal_recall_lr,  fatal_recall_dt,
                       as.numeric(fatal_recall_rf),
                       fatal_recall_xgb, fatal_recall_knn,
                       as.numeric(fatal_recall_nb)), 4),
  Tuned_FR  = round(c(fatal_recall_lr_tuned, fatal_recall_dt_tuned,
                        as.numeric(fatal_recall_tuned),
                        fatal_recall_xgb_tuned, fatal_recall_knn,
                        as.numeric(fatal_recall_nb_tuned)), 4)
)

knitr::kable(
  tuning_compare,
  col.names = c("Model", "Tuning Method",
                "Base Macro-F1", "Tuned Macro-F1",
                "Base Fatal Recall", "Tuned Fatal Recall"),
  caption = "Table 6: Hyperparameter Tuning Summary - Before vs After"
) |>
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "condensed"),
    font_size = 11
  ) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::row_spec(3, bold = TRUE, background = "#D6EAF8") |>
  kableExtra::column_spec(1, bold = TRUE)
```

```{r tuning-curves, echo=TRUE, results='show', fig.width=14, fig.height=9}
plot_grid(p_lr_tune, p_dt_tune, p_rf_tune,
          p_xgb_tune_curve, p_nb_tune, NULL,
          ncol = 3, nrow = 2,
          labels = NULL,
          label_size = 11)
```

**Tuning outcomes by model:**

- **Logistic Regression:** Best decay = `r best_decay`, with the tuning curve showing a clear peak at this value with performance dropping at both lower (no regularisation, overfitting) and higher (over-penalised, underfitting) decay values. Fatal Recall improves marginally while Macro-F1 is essentially unchanged, as LR is already well-regularised by SMOTE inside CV folds
- **Decision Tree:** CV error decreases steadily with more splits before flattening, with best CP = `r signif(best_cp, 3)` selected at the elbow. Pruning prevents the deep tree from over-fitting to the balanced training set, with Macro-F1 improving by +`r round(macro_f1_dt_tuned - macro_f1_dt, 3)` and Fatal Recall by +`r round(fatal_recall_dt_tuned - fatal_recall_dt, 3)`, the largest relative gain among the parametric models
- **Random Forest:** The largest absolute improvement of any model - increasing `mtry` from `r floor(sqrt(ncol(rf_x_train)))` to `r best_mtry` lifted Macro-F1 by +0.007 and pushed Fatal Recall from `r round(as.numeric(fatal_recall_rf), 3)` to `r round(as.numeric(fatal_recall_tuned), 3)`, clearing the 0.50 safety floor. The OOB error curve confirms stability at 100 trees
- **XGBoost:** All six grid combinations cluster tightly in the 0.27-0.30 Macro-F1 range with no setting producing Fatal Recall ≥ 0.50, and depth and learning rate have minimal effect on the structural limitation (poor Fatal class separation under OHE + inverse-frequency weighting)
- **KNN:** No tuning performed, $k = \lfloor\sqrt{p}\rfloor$ rule used (see Section 6.5)
- **Naive Bayes:** Flat log-loss curve across Laplace values confirms the bottleneck is model assumption violation, not hyperparameter choice

------------------------------------------------------------------------

### 7 Model Evaluation & Selection

```{r metric-defs-table, echo=FALSE, results='asis'}
metric_defs <- data.frame(
  Metric = c("Macro-F1", "Fatal Recall", "Weighted F1", "Cohen's Kappa"),
  Definition = c(
    "Unweighted mean of per-class F1 scores",
    "True positive rate for the Fatal class (Sensitivity)",
    "Class-prevalence-weighted mean of per-class F1 scores",
    "Agreement above chance: (Acc - ExpAcc) / (1 - ExpAcc)"
  ),
  Role = c(
    "Primary metric - equal class weighting prevents majority classes masking poor Fatal performance",
    "Hard safety constraint (>= 0.50) - models missing too many Fatal crashes cannot be deployed",
    "Cross-check: penalises misclassification on common classes",
    "Cross-check: robust to class imbalance; near-zero = near-random"
  )
)
knitr::kable(metric_defs, caption = "Table 7: Evaluation Metric Definitions") |>
  kableExtra::kable_styling(full_width = TRUE,
                            bootstrap_options = c("condensed"), font_size = 11) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::column_spec(1, bold = TRUE, width = "16%") |>
  kableExtra::column_spec(2, width = "36%") |>
  kableExtra::column_spec(3, width = "48%")
```

```{r evaluation-table, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
# Build tuned metrics vectors
fr_tuned <- round(c(fatal_recall_lr_tuned, fatal_recall_dt_tuned,
                    as.numeric(fatal_recall_tuned), fatal_recall_xgb_tuned,
                    fatal_recall_knn, as.numeric(fatal_recall_nb_tuned)), 4)
mf1_tuned <- round(c(macro_f1_lr_tuned, macro_f1_dt_tuned,
                     as.numeric(macro_f1_tuned), macro_f1_xgb_tuned,
                     macro_f1_knn, macro_f1_nb_tuned), 4)
passes <- fr_tuned >= 0.50

# Rank passing models by Macro-F1; non-passing get "-"
rank_vec <- rep("-", 6)
passing_idx <- which(passes)
rank_order  <- passing_idx[order(mf1_tuned[passing_idx], decreasing = TRUE)]
rank_vec[rank_order] <- seq_along(rank_order)

decision_vec <- rep("Eliminated (Fatal Recall < 0.50)", 6)
decision_vec[passes & mf1_tuned < 0.50] <- "Pass constraint; low Macro-F1"
decision_vec[passes & mf1_tuned >= 0.50] <- "Dual-pass"
decision_vec[which.max(mf1_tuned * passes)] <- "Selected"

sel_tbl <- data.frame(
  Model         = c("Logistic Regression", "Decision Tree", "Random Forest",
                    "XGBoost", paste0("KNN (k=", k_knn, ")"), "Naive Bayes"),
  `Macro-F1`    = mf1_tuned,
  `Fatal Recall`= fr_tuned,
  `Weighted F1` = round(c(weighted_f1_lr_tuned, weighted_f1_dt_tuned,
                           as.numeric(weighted_f1_tuned), weighted_f1_xgb_tuned,
                           weighted_f1_knn, weighted_f1_nb_tuned), 4),
  Kappa         = round(c(kappa_lr_tuned, kappa_dt_tuned,
                           as.numeric(kappa_tuned), kappa_xgb_tuned,
                           kappa_knn, as.numeric(kappa_nb_tuned)), 4),
  Rank          = rank_vec,
  Decision      = decision_vec,
  check.names   = FALSE
)

selected_row <- which(sel_tbl$Decision == "Selected")

knitr::kable(sel_tbl, align = c("l", "r", "r", "r", "r", "c", "l"),
             caption = "Table 8: Model Selection - Tuned Performance and Decision") |>
  kableExtra::kable_styling(full_width = TRUE,
                            bootstrap_options = c("condensed"), font_size = 12) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::row_spec(selected_row, bold = TRUE, background = "#D5E8D4",
                       color = "#1a5c1a") |>
  kableExtra::column_spec(1, bold = TRUE) |>
  kableExtra::column_spec(3, color = "#d62728") |>
  kableExtra::column_spec(7, bold = TRUE)
```

**Group 1 - Safety floor failure (Fatal Recall < 0.50):**

- **XGBoost** is eliminated: despite grid search over `nrounds`/`max_depth`/`eta`, the model converges to a near-trivial classifier. Kappa near 0 indicates near-random three-class discrimination. Root cause: one-hot expansion of high-cardinality categoricals interacts poorly with the inverse-frequency weighting scheme.

**Group 2 - High-recall but low-F1 (pass constraint, Macro-F1 < 0.50):**

- **Logistic Regression, Decision Tree, and Naive Bayes** all clear the Fatal Recall floor (>= 0.75) but remain below Macro-F1 = 0.50. They over-predict Fatal - high recall at the cost of precision on Injury and Non-casualty.
- **NB** is weakest in this group: its conditional independence assumption is violated by strong correlations between speed limit, urbanisation, lighting, and surface condition, capping both Macro-F1 and Weighted F1.

**Group 3 - Dual-pass models (Fatal Recall >= 0.50 AND Macro-F1 >= 0.50):**

- **KNN** satisfies both constraints but with modest margins (Macro-F1 = `r round(macro_f1_knn, 2)`, Kappa = `r round(kappa_knn, 2)`).
- **Random Forest** dominates across every metric: leads KNN by +`r round(as.numeric(macro_f1_tuned) - macro_f1_knn, 2)` on Macro-F1, +`r round(as.numeric(weighted_f1_tuned) - weighted_f1_knn, 2)` on Weighted F1, and +`r round(as.numeric(kappa_tuned) - kappa_knn, 2)` on Kappa. The ranking is robust to metric choice (class-balanced, prevalence-weighted, or chance-adjusted). **Random Forest is the selected final model.**

**Tuned confusion matrices:**

```{r tuned-cm-grid, echo=FALSE, results='show', fig.width=14, fig.height=9}
plot_grid(p_lr_tuned_cm, p_dt_tuned_cm, p_rf_tuned_cm,
          p_xgb_tuned_cm, p_knn_cm, p_nb_tuned_cm,
          ncol = 3, nrow = 2,
          labels = c("LR", "DT", "RF", "XGBoost", "KNN", "NB"),
          label_size = 11)
```

**Tuned CM observations:**

- **RF (tuned):** The diagonal is now the dominant cell in every row - Fatal, Injury, and Non-casualty predictions each concentrate on the correct class. This is the most balanced CM in the grid and directly reflects the high Macro-F1 (`r round(as.numeric(macro_f1_tuned), 3)`) and cleared Fatal Recall floor (`r round(as.numeric(fatal_recall_tuned), 3)`)
- **KNN:** Similar diagonal structure to RF but with weaker separation on Non-casualty - KNN over-predicts Injury for many Non-casualty instances, explaining the ~`r round(as.numeric(macro_f1_tuned) - macro_f1_knn, 2)` Macro-F1 gap relative to RF
- **LR & DT (tuned):** Tuning made no structural change to the prediction pattern - the Fatal predicted row continues to absorb large proportions of Injury and Non-casualty instances. Both models remain high-recall safety nets, not balanced classifiers
- **XGBoost (tuned):** Predictions are spread roughly equally across all three rows but with very low accuracy in each, with Kappa (`r round(kappa_xgb_tuned, 3)`) confirming near-random three-class discrimination regardless of the predicted class
- **NB (tuned):** High Fatal Recall (`r round(as.numeric(fatal_recall_nb_tuned), 3)`) but the Injury predicted row is severely under-populated - most Injury instances are pushed to Non-casualty. This directly explains the lowest Weighted F1 (`r round(weighted_f1_nb_tuned, 3)`) and Kappa (`r round(as.numeric(kappa_nb_tuned), 3)`) of all tuned models

------------------------------------------------------------------------

### 8 Final Results & Conclusion

```{r holdout-2024, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# 8.1 Prepare 2024 holdout data

year_col <- names(df_model)[grepl("year", names(df_model), ignore.case = TRUE)][1]

df_holdout_2024 <- df_model %>%
  filter(.data[[year_col]] == 2024) %>%
  filter(!is.na(`Degree of crash`))

# RF required variables
required_vars <- names(model_rf_tuned$forest$ncat)

x_2024 <- df_holdout_2024

# Add missing columns if needed
missing_vars <- setdiff(required_vars, names(x_2024))

for (v in missing_vars) {
  x_2024[[v]] <- NA
}

# Keep same variables and order
x_2024 <- x_2024[, required_vars, drop = FALSE]

# Pre-compute training set medians for numeric variables (use train, not holdout, to avoid leakage)
train_medians <- sapply(
  rf_x_train[, sapply(rf_x_train, is.numeric), drop = FALSE],
  median, na.rm = TRUE
)

# Handle missing values and factor levels
for (v in required_vars) {

  # If variable is categorical in RF
  if (v %in% names(model_rf_tuned$forest$xlevels)) {

    train_levels <- model_rf_tuned$forest$xlevels[[v]]

    # Convert to character first
    x_2024[[v]] <- as.character(x_2024[[v]])

    # Replace unseen or missing levels
    replacement_level <- if ("Unknown" %in% train_levels) {
      "Unknown"
    } else if ("Other" %in% train_levels) {
      "Other"
    } else {
      train_levels[1]
    }

    x_2024[[v]][is.na(x_2024[[v]]) | !(x_2024[[v]] %in% train_levels)] <- replacement_level

    x_2024[[v]] <- factor(x_2024[[v]], levels = train_levels)

  } else {

    # Numeric: impute with training median, not holdout median
    x_2024[[v]] <- as.numeric(x_2024[[v]])
    fill_val <- if (!is.null(train_medians[[v]]) && !is.na(train_medians[[v]])) {
      train_medians[[v]]
    } else {
      0
    }
    x_2024[[v]][is.na(x_2024[[v]])] <- fill_val
  }
}

# Outcome
y_2024 <- factor(
  df_holdout_2024$`Degree of crash`,
  levels = c("Fatal", "Injury", "Non_casualty")
)

# Check row count
nrow(x_2024)
table(y_2024)
# 8.2 Predict on 2024 holdout

pred_rf_2024 <- predict(

  model_rf_tuned,

  newdata = x_2024

)

cm_rf_2024 <- confusionMatrix(

  pred_rf_2024,

  y_2024

)

cm_rf_2024
# 8.3 Compute holdout metrics

recall_2024 <- cm_rf_2024$byClass[, "Sensitivity"]
precision_2024 <- cm_rf_2024$byClass[, "Pos Pred Value"]

f1_2024 <- 2 * precision_2024 * recall_2024 /
  (precision_2024 + recall_2024)

# Treat undefined F1 as 0
f1_2024[is.na(f1_2024)] <- 0

macro_f1_2024 <- mean(f1_2024)

fatal_recall_2024 <- recall_2024["Class: Fatal"]

class_weights_2024 <- prop.table(table(y_2024))
f1_names_2024 <- gsub("Class: ", "", names(f1_2024))

weighted_f1_2024 <- sum(
  f1_2024 * class_weights_2024[f1_names_2024],
  na.rm = TRUE
)

kappa_2024 <- cm_rf_2024$overall["Kappa"]

rf_2024_metrics <- data.frame(
  Model = "Tuned RF - 2024 Holdout",
  Macro_F1 = round(macro_f1_2024, 3),
  Fatal_Recall = round(as.numeric(fatal_recall_2024), 3),
  Weighted_F1 = round(weighted_f1_2024, 3),
  Kappa = round(as.numeric(kappa_2024), 3),
  row.names = NULL
)

knitr::kable(
  rf_2024_metrics,
  caption = "Tuned Random Forest Performance on 2024 Holdout Data"
)
# 8.4 Holdout confusion matrix

cm_rf_2024_df <- as.data.frame(cm_rf_2024$table)

p_rf_2024_cm <- ggplot(
  cm_rf_2024_df,
  aes(x = Reference, y = Prediction, fill = Freq)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "RF (Tuned) - 2024 Holdout Confusion Matrix",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_2024, 3),
                      "  |  Macro-F1: ", round(macro_f1_2024, 3)),
    x = "Actual", y = "Predicted",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 10))

p_rf_2024_cm

ggsave(
  "plots/8_rf_2024_holdout_confusion_matrix.png",
  p_rf_2024_cm,
  width = 6,
  height = 4,
  dpi = 220
)
```

```{r holdout-display, echo=TRUE, results='asis'}
knitr::kable(
  rf_2024_metrics,
  col.names = c("Model", "Macro-F1", "Fatal Recall", "Weighted F1", "Kappa"),
  caption = "Table 9: Tuned Random Forest - 2024 Temporal Holdout Performance"
) |>
  kableExtra::kable_styling(full_width = FALSE,
                            bootstrap_options = c("striped", "condensed"),
                            font_size = 12) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::column_spec(1, bold = TRUE)
```

**Model selection:** Tuned Random Forest was selected as the final model based on test-set performance (Macro-F1 = `r round(as.numeric(macro_f1_tuned), 3)`, Fatal Recall = `r round(as.numeric(fatal_recall_tuned), 3)`). It was the only model to satisfy both the Macro-F1 ≥ 0.50 and Fatal Recall ≥ 0.50 constraints simultaneously, and led the next-best model (KNN) by +`r round(as.numeric(macro_f1_tuned) - macro_f1_knn, 2)` on Macro-F1 and +`r round(as.numeric(kappa_tuned) - kappa_knn, 2)` on Kappa. The holdout confusion matrix and top-15 variable importance are shown below.

```{r holdout-cm-varimp, echo=TRUE, results='show', fig.width=13, fig.height=5}
imp_df_final <- as.data.frame(importance(model_rf_tuned, type = 1))
imp_df_final$Feature <- rownames(imp_df_final)
p_varimp_final <- imp_df_final %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(Feature, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_col(fill = "#2E74B5") +
  coord_flip() +
  labs(title = "Variable Importance (Top 15)",
       x = NULL, y = "Mean Decrease in Accuracy") +
  theme_minimal(base_size = 11)

plot_grid(p_rf_2024_cm, p_varimp_final, ncol = 2, rel_widths = c(1, 1.2))
```

**Holdout performance (2024 unseen data):**

- Fatal Recall collapsed from `r round(as.numeric(fatal_recall_tuned), 3)` (test set) to `r round(as.numeric(fatal_recall_2024), 3)` - only `r round(as.numeric(fatal_recall_2024) * 100, 1)`% of actual Fatal crashes correctly identified, with the rest misclassified as Injury
- Macro-F1 fell from `r round(as.numeric(macro_f1_tuned), 3)` to `r round(macro_f1_2024, 3)`, with Kappa dropping to `r round(as.numeric(kappa_2024), 3)` (near-chance agreement)
- Weighted F1 (`r round(weighted_f1_2024, 3)`) is deceptively high - the model predicts Injury for almost everything, and Injury dominates the 2024 holdout

**Why generalisation failed:**

- **Temporal distribution shift** - training used a balanced set (1:10:10 ratio) but 2024 holdout reflects the natural ~1:47:23 distribution, and decision boundaries trained on the balanced set do not transfer
- **Top predictors are time-sensitive** - Street lighting, LGA, and Key TU type shift with infrastructure upgrades and reporting practice changes year-on-year
- **No recalibration** - predicted probabilities were not threshold-calibrated for the Fatal class, with the default 0.33 threshold being too high given 1.4% natural prevalence

**Variable importance findings:**

- Street lighting and LGA dominate importance, suggesting crash severity in NSW is strongly tied to infrastructure quality and geographic context
- Temporal features (Month, Day of week, Two-hour intervals) appear in the top 15, confirming time-of-day and seasonal patterns carry predictive signal
- Speed limit ranks lower than expected - likely because most crashes cluster in 50-60 km/h zones, reducing its discriminative power

**Limitations:**

- Undersampling discards ~80% of majority-class rows, reducing the information available for learning Injury/Non-casualty boundaries
- Euclidean distance on OHE features (used by KNN) is a known limitation for high-cardinality categorical data
- Model was trained on a static 2020-2023 window with no temporal trend features

**Future work:**

- Rolling-window retraining (e.g., retrain annually on most recent 3 years) to track distribution drift
- Probability threshold calibration for the Fatal class to recover recall under natural class prevalence
- Explore crash-level aggregation strategies and time-series features to explicitly model year-on-year trends

**Answer to the research question:**

Given environmental conditions and traffic unit characteristics, can we predict NSW road crash severity as a three-class problem (Fatal / Injury / Non-casualty)? **Partially yes.** The tuned Random Forest achieves strong discrimination on the 2020-2023 test set (Macro-F1 = `r round(as.numeric(macro_f1_tuned), 3)`, Fatal Recall = `r round(as.numeric(fatal_recall_tuned), 3)`), demonstrating that the chosen predictor set carries genuine signal for separating severity classes. However, the sharp generalisation drop on the 2024 holdout (Fatal Recall = `r round(as.numeric(fatal_recall_2024), 3)`) shows that prediction is sensitive to temporal distribution shift and cannot yet be considered reliable for deployment without periodic retraining and threshold recalibration.

------------------------------------------------------------------------

#### References

<p style="font-size:14px;">Chawla, N.V., Bowyer, K.W., Hall, L.O., & Kegelmeyer, W.P. (2002). SMOTE: Synthetic Minority Over-sampling Technique. *Journal of Artificial Intelligence Research*, 16, 321–357. <a href="https://www.jair.org/index.php/jair/article/view/10302">https://doi.org/10.1613/jair.953</a></p>

<p style="font-size:14px;">Cover, T.M. & Hart, P.E. (1967). Nearest neighbor pattern classification. *IEEE Transactions on Information Theory*, 13(1), 21–27. <a href="https://ieeexplore.ieee.org/document/1053964">https://doi.org/10.1109/TIT.1967.1053964</a></p>

<p style="font-size:14px;">Hastie, T., Tibshirani, R., & Friedman, J. (2009). *The Elements of Statistical Learning* (2nd ed.). Springer. <a href="https://hastie.su.domains/ElemStatLearn/">https://hastie.su.domains/ElemStatLearn/</a></p>

<p style="font-size:14px;">James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning* (2nd ed.). Springer. <a href="https://www.statlearning.com">https://www.statlearning.com</a></p>

<p style="font-size:14px;">Transport for NSW. (2024). NSW road crash data 2020–2024 [Dataset]. Australian Government Open Data Platform. https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data</p>

------------------------------------------------------------------------

#### Acknowledgement of AI

<p style="font-size:14px;">Group 07 (Workshop 12) declares that we used ChatGPT (GPT-5.3, OpenAI, https://chatgpt.com) and Claude Sonnet 4.5 (Anthropic, https://claude.ai) for code refinement, grammar checking, making informed coding decisions, and reviewing report content against the marking rubric. The authors retain full understanding of all decisions made with AI assistance.</p>
