---
title: "NSW Road Crash Severity Prediction"
subtitle: "STAT5003  -  Group 07, Workshop 12"
author: "Sanskar | Robert | Haipeng | Chen | Yihan"
date: "May 2026"
output:
  html_document:
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  results = 'hide',
  warning = FALSE,
  message = FALSE
)
```

```{r library, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(kableExtra)
library(caret)
library(nnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(kknn)
library(e1071)
library(themis)
library(pROC)
library(flextable)
library(MLmetrics)
library(Matrix)
library(tibble)
library(naivebayes)
library(cowplot)
library(gridExtra)

# Create plots output folder  -  all ggsave() calls write here for use in Presentation Rmd
dir.create("plots", showWarnings = FALSE)
```

```{css, echo=FALSE}
body { font-family: "Times New Roman", Times, serif; font-size: 10pt; line-height: 1.3; }
h2 { margin-top: 0.4em; margin-bottom: 0.15em; }
h3 { margin-top: 0.3em; margin-bottom: 0.1em; }
h4 { margin-top: 0.25em; margin-bottom: 0.1em; }
p  { margin-top: 0.2em; margin-bottom: 0.2em; }
h1.title { display: none; }
h4.author { display: none; }
h4.date   { display: none; }
```

```{r load-data, message=FALSE, warning=FALSE, results='hide'}
df <- read.csv("nsw_road_crash_merged_2020_2024.csv", check.names = FALSE)
cat("Dimensions:", nrow(df), "rows x", ncol(df), "cols\n")
str(df)
```

------------------------------------------------------------------------

### 1 Problem Framing & Research Question

#### 1.1 Research Question
Given environmental conditions (weather, natural lighting, road surface, speed limit) and traffic unit characteristics (type of traffic unit, manoeuvre, number of units involved), can we predict the severity of a road crash in NSW as a three-class classification problem: Fatal, Injury, or Non-casualty?

#### 1.2 Motivation
NSW recorded 340 road fatalities in 2024, with over two-thirds occurring in regional areas (Transport for NSW, 2025), highlighting the need for data-driven approaches to understanding crash severity. Fatal crashes represent only approximately 1.4% of all crashes, introducing severe class imbalance that must be carefully addressed during modelling and evaluation.

A predictive model for crash severity has direct real-world value for the following stakeholders:
NSW Traffic Management Authorities: By identifying key combinations of environmental and traffic characteristics associated with severe crashes, authorities can make targeted decisions such as adjusting speed limits and improving road design in high-risk areas.

Emergency Response Teams: Predicting crash severity based on environmental conditions in the early stages of an incident can support faster and more evidence-based allocation of ambulance and emergency resources.
Policymakers and Insurance Institutions: Understanding which traffic unit types and environmental conditions are associated with higher severity risk can inform more precise road safety policies and insurance pricing models.

------------------------------------------------------------------------

### 2 Dataset & Justification

#### 2.1 Dataset Overview

```{r dataset-overview, echo=FALSE, message=FALSE, warning=FALSE}
overview <- data.frame(
  Item = c(
    "Time Period",
    "Data Sources",
    "Merge Key",
    "Total Records",
    "Total Variables",
    "Target Variable",
    "Fatal",
    "Injury",
    "Non-casualty"
  ),
  Detail = c(
    "2020 - 2024",
    "CRASH.xlsx + TRAFFIC UNIT.xlsx (TfNSW via data.gov.au)",
    "Crash ID (left join)",
    format(nrow(df), big.mark = ","),
    as.character(ncol(df)),
    "Degree of Crash (3-class classification)",
    paste0(format(sum(df$`Degree of crash` == "Fatal"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Fatal") * 100, 1), "%)"),
    paste0(format(sum(df$`Degree of crash` == "Injury"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Injury") * 100, 1), "%)"),
    paste0(format(sum(df$`Degree of crash` == "Non-casualty (towaway)"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Non-casualty (towaway)") * 100, 1), "%)")
  )
)

knitr::kable(overview, col.names = c("Item", "Detail"),
             caption = "Table 1: Dataset Overview")
```

------------------------------------------------------------------------

### 3 EDA & Insights
```{r 311-selected, fig.width=12, fig.height=4, message=FALSE, warning=FALSE}
library(gridExtra)

p1 <- ggplot(df, aes(x = fct_infreq(`Degree of crash`))) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3) +
    labs(title = "Distribution of Crash Severity", x = "Severity", y = "Count") +
    theme_minimal() + theme(legend.position = "none")

p2 <- ggplot(df, aes(x = `Speed limit`)) +
  geom_bar(fill = "tomato") +
  labs(title = "Distribution of Speed Limit", x = "Speed Limit", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 <- ggplot(df, aes(x = fct_infreq(`TU type group`))) +
    geom_bar(fill = "steelblue") +
    labs(title = "Distribution of TU Type", x = "TU Type", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4 <- ggplot(df, aes(x = fct_infreq(Urbanisation))) +
  geom_bar(fill = "purple") +
  labs(title = "Distribution of Urbanisation", x = "Urbanisation", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1, p2, p3, p4, nrow = 1)
```

- **Crash Severity:** Fatal crashes represent only 1.4% of records, indicating severe class imbalance that must be addressed during modelling.
- **Speed Limit:** Crashes are heavily concentrated in 50-60 km/h urban zones, reflecting higher traffic volume rather than elevated per-trip risk.
- **Traffic Unit Type:** Car/car derivatives dominate (~114,477), with pedestrians and motorcycles being rare but disproportionately associated with severe outcomes.
- **Urbanisation:** Sydney metro accounts for the most crashes (98,975), while Country non-urban regions show higher fatal proportions despite lower overall volume.

------------------------------------------------------------------------

### 4 Data Preprocessing

All data-dependent steps are performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

```{r preprocessing-all, echo=TRUE, results='hide'}
# 4.1 Data Integration
# Each row is one traffic unit - summarise binary flags per crash, then join back
tu_flags <- df %>%
  group_by(`Crash ID`) %>%
  summarise(
    has_motorcycle    = as.integer(any(`TU type group` == "Motorcycle")),
    has_heavy_vehicle = as.integer(any(`TU type group` %in% c("Articulated truck", "Heavy rigid truck", "Bus"))),
    has_pedestrian    = as.integer(any(`TU type group` == "Pedestrian")),
    has_cyclist       = as.integer(any(`TU type group` == "Pedal cycle")),
    tu_count          = n(),
    .groups = "drop"
  )
df <- df %>% left_join(tu_flags, by = "Crash ID")

# 4.2 Feature Removal
leakage_cols <- c("No. killed", "No. seriously injured",
                  "No. moderately injured", "No. minor-other injured")
redundant_target_cols <- c("Degree of crash - detailed")
id_cols <- c("Crash ID", "Traffic unit ID", "Street of crash",
             "Street of travel", "Town", "Identifying feature",
             "Identifying feature type", "Conurbation 1")
high_na_cols <- c("Primary hazardous feature", "Primary temporary feature",
                  "Object hit 2", "DCA supplement", "Object hit 1",
                  "Primary permanent feature")
redundant_cols <- c("Route no.", "RUM - code", "RUM - description",
                    "DCA - code", "DCA - description")
drop_cols <- unique(c(leakage_cols, redundant_target_cols, id_cols, high_na_cols, redundant_cols))
df_clean  <- df %>% select(-any_of(drop_cols))

# 4.3 Missing Value Imputation
df_clean <- df_clean %>%
  mutate(
    `Other TU type` = ifelse(is.na(`Other TU type`), "Unknown", `Other TU type`),
    across(where(is.character), ~ replace_na(., "Unknown")),
    across(where(is.numeric),   ~ replace_na(., median(., na.rm = TRUE)))
  )

# 4.4 Feature Engineering
df_clean <- df_clean %>%
  mutate(
    crash_hour   = as.integer(sub(":.*", "", `Two-hour intervals`)),
    is_weekend   = as.integer(`Day of week of crash` %in% c("Saturday", "Sunday")),
    is_rush_hour = as.integer(crash_hour %in% c(7, 8, 9, 16, 17, 18)),
    is_night     = as.integer(`Natural lighting` %in% c("Darkness", "Dawn", "Dusk")),
    is_wet       = as.integer(`Surface condition` %in% c("Wet", "Snow or ice")),
    speed_band   = case_when(
      `Speed limit` %in% c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h") ~ "low_50",
      `Speed limit` %in% c("60 km/h","70 km/h","80 km/h")                     ~ "mid_60_80",
      `Speed limit` %in% c("90 km/h","100 km/h")                              ~ "high_90_100",
      `Speed limit` == "110 km/h"                                              ~ "very_high_110",
      TRUE                                                                     ~ "unknown"
    )
  )

# 4.5 Outlier Handling
p99_dist <- quantile(df_clean$Distance, 0.99, na.rm = TRUE)
df_clean <- df_clean %>%
  mutate(Distance = pmin(Distance, p99_dist)) %>%
  filter(is.na(Latitude)  | (Latitude  >= -37.5 & Latitude  <= -28.2),
         is.na(Longitude) | (Longitude >= 141.0  & Longitude <= 153.6))

# 4.6 Encoding
lga_freq  <- prop.table(table(df_clean$LGA))
rare_lgas <- names(lga_freq[lga_freq < 0.01])
df_clean  <- df_clean %>%
  mutate(LGA = ifelse(LGA %in% rare_lgas, "Other", LGA))
df_model <- df_clean %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(`Degree of crash` = factor(`Degree of crash`,
           levels = c("Fatal", "Injury", "Non-casualty (towaway)"),
           labels = c("Fatal", "Injury", "Non_casualty")))

# 4.7 Scaling - deferred to per-model preprocessing (KNN uses caret preProcess = c("center","scale"))

# 4.8 Class Imbalance - 2024 sealed as temporal holdout; undersample Injury/Non-casualty to 15k each
df_holdout_2024  <- df_model %>% filter(`Reporting year` == 2024)
df_model_2020_23 <- df_model %>% filter(`Reporting year` != 2024)
set.seed(42)
train_idx <- createDataPartition(df_model_2020_23$`Degree of crash`, p = 0.9, list = FALSE)
df_train  <- df_model_2020_23[ train_idx, ]
df_test   <- df_model_2020_23[-train_idx, ]

set.seed(42)
fatal_rows   <- df_train %>% filter(`Degree of crash` == "Fatal")
injury_rows  <- df_train %>% filter(`Degree of crash` == "Injury") %>% slice_sample(n = 15000)
noncas_rows  <- df_train %>% filter(`Degree of crash` == "Non_casualty") %>% slice_sample(n = 15000)
df_train_bal <- bind_rows(fatal_rows, injury_rows, noncas_rows) %>% slice_sample(prop = 1)

# Inverse frequency weights
class_weights <- 1 / prop.table(table(df_train_bal$`Degree of crash`))
class_weights <- class_weights / sum(class_weights)

# CV controls (SMOTE applied inside folds for LR and NB only)
cv_ctrl <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final"
)
cv_ctrl_smote <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final",
  sampling = "smote"
)

# 4.9 Train/Test Split
cat("Train (balanced):", nrow(df_train_bal),
    "| Test:", nrow(df_test),
    "| Holdout (2024):", nrow(df_holdout_2024), "\n")
cat("df_train_bal class distribution:\n")
print(prop.table(table(df_train_bal$`Degree of crash`)))
```

Nine sequential preprocessing steps convert the raw merged dataset into `df_train_bal`, the balanced training frame used by all six models. Traffic-unit rows were aggregated into per-crash binary flags (`has_motorcycle`, `has_pedestrian`, etc.) to avoid row duplication from the left join. Post-crash injury counts were dropped as leakage, along with identifiers, free-text fields, and columns exceeding 50% missingness. Remaining NAs were imputed with `"Unknown"` for categoricals and the column median for numerics — median was chosen over mean because several numeric predictors (e.g., `Distance`) are right-skewed.

Five contextual features were engineered from existing columns: `is_weekend`, `is_rush_hour`, `is_night`, `is_wet`, and `speed_band`. Distance was winsorised at the 99th percentile and GPS coordinates outside NSW bounds were removed. LGA levels below 1% frequency were pooled into `"Other"` to reduce cardinality before factor encoding.

The original Fatal:Injury:Non-casualty ratio of ~1:47:23 was far too imbalanced for reliable Fatal detection. Random undersampling was applied: all ~1,500 Fatal rows were retained while Injury and Non-casualty were each capped at 15,000, producing `df_train_bal` (~31,500 rows, ratio ~1:10:10). Undersampling was chosen over pure SMOTE because the tiny Fatal class (~770 rows after the 90/10 split) provides an unstable base for synthetic generation. Inverse-frequency class weights were added for tree-based models; SMOTE was applied inside CV folds for Logistic Regression and Naive Bayes only.

Injury and Non-casualty were capped equally at 15,000 rather than sampled proportionally to their original ~47:23 ratio for two reasons. First, proportional sampling would preserve ~36,000 Injury rows against ~18,000 Non-casualty — still a 2:1 imbalance between the two majority classes, causing the model to learn the Injury/Non-casualty boundary unevenly. Equal capping ensures both majority classes are represented with comparable fidelity. Second, 15,000 per class was chosen to capture sufficient within-class diversity (each class spans varied road types, lighting, and traffic unit combinations) while keeping the total training set computationally tractable (~31,500 rows) within a single knit session.

<!-- ------------------------------------------------------------------------ -->

### 5 Model Fitting

Six classifiers were fitted to cover a range of complexity and interpretability — from a parametric baseline (Logistic Regression) through tree ensembles (Decision Tree, Random Forest, XGBoost) to distance- and probability-based models (KNN, Naive Bayes). All models train on the same `df_train_bal` (~31,500 rows) and are evaluated on the same held-out test set; cross-model differences therefore reflect genuine algorithmic variation rather than data preparation differences.

```{r models-all, echo=TRUE, results='hide', fig.show='hide'}
# 5.1 Logistic Regression - baseline fit; class imbalance handled via SMOTE inside CV folds
set.seed(42)

train_lr <- df_train_bal %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

test_lr <- df_test %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

model_lr <- train(
  Degree.of.crash ~ .,
  data = train_lr,
  method = "multinom",
  trControl = cv_ctrl_smote,
  preProcess = "nzv",
  metric = "Mean_F1",
  trace = FALSE,
  MaxNWts = 20000
)

print(model_lr$results)

pred_lr <- predict(model_lr, newdata = test_lr)
cm_lr   <- confusionMatrix(pred_lr, test_lr$Degree.of.crash)

fatal_recall_lr <- as.numeric(cm_lr$byClass["Class: Fatal", "Recall"])
f1_lr_vec <- setNames(cm_lr$byClass[, "F1"], gsub("Class: ", "", rownames(cm_lr$byClass)))
test_props_lr <- prop.table(table(test_lr$Degree.of.crash))
macro_f1_lr <- mean(f1_lr_vec, na.rm = TRUE)
weighted_f1_lr <- weighted.mean(f1_lr_vec[names(test_props_lr)], test_props_lr, na.rm = TRUE)
kappa_lr <- as.numeric(cm_lr$overall["Kappa"])

cat(sprintf(
  "   Logistic Regression  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_lr, fatal_recall_lr, weighted_f1_lr, kappa_lr
))

# Confusion matrix heatmap
p_lr_cm <- as.data.frame(cm_lr$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Logistic Regression  -  Confusion Matrix (Test Set)",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_lr, 3),
                      "  |  Macro-F1: ", round(macro_f1_lr, 3))
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_lr_cm)
ggsave("plots/5_1_logistic_regression_confusion_matrix.png", p_lr_cm, width = 7, height = 5, dpi = 150)

# Coefficient plot  -  top 15 absolute coefficients across non-reference classes
coef_lr <- coef(model_lr$finalModel)

if (is.null(dim(coef_lr))) {
  coef_lr <- matrix(coef_lr, nrow = 1)
  rownames(coef_lr) <- "Non-reference class"
}

coef_lr_df <- as.data.frame(as.table(coef_lr))
if (ncol(coef_lr_df) == 3) {
  names(coef_lr_df) <- c("Class", "Feature", "Coefficient")
} else if (ncol(coef_lr_df) == 2) {
  names(coef_lr_df) <- c("Feature", "Coefficient")
  coef_lr_df$Class <- "Non-reference class"
  coef_lr_df <- coef_lr_df %>% select(Class, Feature, Coefficient)
}

coef_lr_df <- coef_lr_df %>%
  filter(Feature != "(Intercept)") %>%
  mutate(abs_coef = abs(as.numeric(Coefficient))) %>%
  arrange(desc(abs_coef)) %>%
  slice_head(n = 15)

p_lr_coef <- ggplot(coef_lr_df, aes(x = reorder(Feature, abs_coef), y = as.numeric(Coefficient), fill = Class)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Logistic Regression  -  Largest Coefficients (Top 15)",
    x = NULL,
    y = "Coefficient (log-odds)"
  ) +
  theme_minimal(base_size = 12)
print(p_lr_coef)
ggsave("plots/5_1_logistic_regression_coefficients.png", p_lr_coef, width = 7, height = 5, dpi = 150)

# 5.2 Decision Tree - direct rpart(); OHE via dummyVars; caret formula interface unstable here
set.seed(42)

# 1. First standardize column names with `make.names` to prevent `dummyVars` from failing to locate columns due to spaces.
train_dt_clean <- df_train_bal %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

test_dt_clean <- df_test %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

# 2. Establish One-Hot encoding
dummy_model <- dummyVars(Degree.of.crash ~ ., data = train_dt_clean)

# 3. Transform the training set
train_dt_x <- as.data.frame(predict(dummy_model, newdata = train_dt_clean))
train_dt <- cbind(train_dt_x, Degree.of.crash = train_dt_clean$Degree.of.crash)

# 4. Transform the test set and concatenate back the target variable
test_dt_x <- as.data.frame(predict(dummy_model, newdata = test_dt_clean))
test_dt <- cbind(test_dt_x, Degree.of.crash = test_dt_clean$Degree.of.crash)

# 5. Align weights
dt_weights <- as.numeric(class_weights[as.character(train_dt$Degree.of.crash)])

# 6. Train the decision tree
model_dt <- rpart(
  Degree.of.crash ~ .,
  data = train_dt,
  method = "class",
  weights = dt_weights,
  control = rpart.control(cp = 0.01, xval = 10)
)

pred_dt <- predict(model_dt, newdata = test_dt, type = "class")
cm_dt   <- confusionMatrix(pred_dt, test_dt$Degree.of.crash)

fatal_recall_dt <- as.numeric(cm_dt$byClass["Class: Fatal", "Recall"])
f1_dt_vec <- setNames(cm_dt$byClass[, "F1"], gsub("Class: ", "", rownames(cm_dt$byClass)))
test_props_dt <- prop.table(table(test_dt$Degree.of.crash))
macro_f1_dt <- mean(f1_dt_vec, na.rm = TRUE)
weighted_f1_dt <- weighted.mean(f1_dt_vec[names(test_props_dt)], test_props_dt, na.rm = TRUE)
kappa_dt <- as.numeric(cm_dt$overall["Kappa"])

cat(sprintf(
  "   Decision Tree  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_dt, fatal_recall_dt, weighted_f1_dt, kappa_dt
))

printcp(model_dt)

# Tree diagram  -  use a simpler, more robust plotting configuration
png("plots/5_2_decision_tree_diagram.png", width = 1200, height = 800, res = 150)
rpart.plot(
  model_dt,
  type = 2,
  extra = 1,
  under = TRUE,
  fallen.leaves = TRUE,
  cex = 0.7,
  main = "Decision Tree Diagram"
)
invisible(dev.off())

rpart.plot(
  model_dt,
  type = 2,
  extra = 1,
  under = TRUE,
  fallen.leaves = TRUE,
  cex = 0.7,
  main = "Decision Tree Diagram"
)

# Confusion matrix heatmap
p_dt_cm <- as.data.frame(cm_dt$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Decision Tree  -  Confusion Matrix (Test Set)",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_dt, 3),
                      "  |  Macro-F1: ", round(macro_f1_dt, 3))
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_dt_cm)
ggsave("plots/5_2_decision_tree_confusion_matrix.png", p_dt_cm, width = 7, height = 5, dpi = 150)

# 5.3 Random Forest - randomForest() direct; OOB used as CV estimate; na.roughfix() for residual NAs
rf_x_train <- na.roughfix(df_train_bal %>%
  select(-`Degree of crash`) %>%
  select(where(~ !is.factor(.) || nlevels(.) <= 53)) %>%
  droplevels())

# Align test factor levels to training  -  unseen levels become NA, then imputed
rf_x_test_raw <- df_test %>% select(all_of(names(rf_x_train)))
for (col in names(rf_x_test_raw)) {
  if (is.factor(rf_x_test_raw[[col]]))
    rf_x_test_raw[[col]] <- factor(rf_x_test_raw[[col]], levels = levels(rf_x_train[[col]]))
}
rf_x_test <- na.roughfix(rf_x_test_raw)

set.seed(42)
model_rf <- randomForest(
  x          = rf_x_train,
  y          = df_train_bal$`Degree of crash`,
  mtry       = floor(sqrt(ncol(rf_x_train))),
  ntree      = 100,
  importance = TRUE
)

# Test set predictions and metrics
pred_rf         <- predict(model_rf, rf_x_test)
cm_rf           <- confusionMatrix(pred_rf, df_test$`Degree of crash`)
fatal_recall_rf <- cm_rf$byClass["Class: Fatal", "Recall"]
macro_f1_rf     <- mean(cm_rf$byClass[, "F1"],  na.rm = TRUE)
weighted_f1_rf  <- weighted.mean(cm_rf$byClass[, "F1"],
                    prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_rf        <- cm_rf$overall["Kappa"]

cat(sprintf(
  "   Random Forest  -   Test Set Metrics\nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf$err.rate[100, "OOB"], macro_f1_rf, fatal_recall_rf, weighted_f1_rf, kappa_rf
))

# Confusion matrix heatmap
p_rf_cm <- as.data.frame(cm_rf$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(title = "Random Forest : Confusion Matrix (Test Set)",
       subtitle = paste0("Fatal Recall: ", round(fatal_recall_rf, 3),
                         "  |  Macro-F1: ", round(macro_f1_rf, 3))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_rf_cm)
ggsave("plots/5_3_random_forest_confusion_matrix.png", p_rf_cm, width = 7, height = 5, dpi = 150)

# Variable importance (top 15) : MeanDecreaseAccuracy across all classes
imp_df <- as.data.frame(importance(model_rf, type = 1))
imp_df$Feature <- rownames(imp_df)
p_rf_varimp <- imp_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(Feature, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_col(fill = "#2E74B5") +
  coord_flip() +
  labs(title = "Random Forest : Variable Importance (Top 15)",
       x = NULL, y = "Mean Decrease in Accuracy") +
  theme_minimal(base_size = 12)
print(p_rf_varimp)
ggsave("plots/5_3_random_forest_variable_importance_top15.png", p_rf_varimp, width = 7, height = 5, dpi = 150)

# 5.4 XGBoost - OHE via model.matrix(); class weights via obs_weights; defaults: depth=6, eta=0.3, rounds=100

target_col <- "Degree of crash"

x_train_raw <- df_train_bal %>% select(-all_of(target_col))

x_test_raw <- df_test %>% select(-all_of(target_col))

y_train_factor <- df_train_bal[[target_col]]

y_test_factor  <- df_test[[target_col]]

y_train_factor <- factor(
  y_train_factor,
  levels = c("Fatal", "Injury", "Non_casualty")
)

y_test_factor <- factor(
  y_test_factor,
  levels = levels(y_train_factor)
)

class_levels <- levels(y_train_factor)

# XGBoost requires labels as 0, 1, 2

y_train <- as.integer(y_train_factor) - 1

y_test  <- as.integer(y_test_factor) - 1

table(y_train_factor)

train_crash_hour_median <- median(x_train_raw$crash_hour, na.rm = TRUE)

x_train_raw <- x_train_raw %>% mutate(
    crash_hour = ifelse(
      is.na(crash_hour),
      train_crash_hour_median,
      crash_hour
    )
  )

x_test_raw <- x_test_raw %>%
  mutate(
    crash_hour = ifelse(
      is.na(crash_hour),
      train_crash_hour_median,
      crash_hour
    )
  )

x_train_mm <- model.matrix(~ . - 1, data = x_train_raw)

x_test_mm  <- model.matrix(~ . - 1, data = x_test_raw)

# Clean feature names for XGBoost compatibility

colnames(x_train_mm) <- make.names(colnames(x_train_mm), unique = TRUE)

colnames(x_test_mm)  <- make.names(colnames(x_test_mm), unique = TRUE)

# Align test columns with training columns

missing_cols <- setdiff(colnames(x_train_mm), colnames(x_test_mm))

if (length(missing_cols) > 0) {
  missing_matrix <- matrix(
    0,
    nrow = nrow(x_test_mm),
    ncol = length(missing_cols)
  )
  colnames(missing_matrix) <- missing_cols
  x_test_mm <- cbind(x_test_mm, missing_matrix)

}

extra_cols <- setdiff(colnames(x_test_mm), colnames(x_train_mm))

if (length(extra_cols) > 0) {
  x_test_mm <- x_test_mm[, !(colnames(x_test_mm) %in% extra_cols), drop = FALSE]
}

x_test_mm <- x_test_mm[, colnames(x_train_mm), drop = FALSE]

# Final checks

stopifnot(nrow(x_train_mm) == length(y_train))

stopifnot(nrow(x_test_mm) == length(y_test))

stopifnot(identical(colnames(x_train_mm), colnames(x_test_mm)))

stopifnot(!any(is.na(x_train_mm)))

stopifnot(!any(is.na(x_test_mm)))

class_counts <- table(y_train_factor)

class_weights <- sum(class_counts) /(length(class_counts) * class_counts)

obs_weights <- as.numeric(class_weights[y_train_factor])

stopifnot(length(obs_weights) == nrow(x_train_mm))

stopifnot(!any(is.na(obs_weights)))

stopifnot(!any(is.infinite(obs_weights)))

dtrain <- xgb.DMatrix(
  data = x_train_mm,
  label = y_train,
  weight = obs_weights
)

dtest <- xgb.DMatrix(
  data = x_test_mm,
  label = y_test
)

xgb_params <- list(
  objective = "multi:softprob",
  num_class = length(class_levels),
  eval_metric = "mlogloss",
  max_depth = 6,
  eta = 0.3,
  gamma = 0,
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = 1
)

set.seed(5003)

xgb_fit <- xgb.train(
  params = xgb_params,
  data = dtrain,
  nrounds = 100,
  verbose = 0
)

xgb_prob <- predict(xgb_fit, dtest)

xgb_prob_matrix <- matrix(
  xgb_prob,
  ncol = length(class_levels),
  byrow = TRUE
)

colnames(xgb_prob_matrix) <- class_levels

xgb_pred_index <- max.col(xgb_prob_matrix) - 1

xgb_pred <- factor(
  class_levels[xgb_pred_index + 1],
  levels = class_levels
)

cm_xgb <- confusionMatrix(
  data = xgb_pred,
  reference = y_test_factor
)

accuracy_xgb <- as.numeric(cm_xgb$overall["Accuracy"])

kappa_xgb    <- as.numeric(cm_xgb$overall["Kappa"])

by_class_xgb <- as.data.frame(cm_xgb$byClass) %>% rownames_to_column("Class")

macro_f1_xgb <- mean(
  by_class_xgb$F1,
  na.rm = TRUE
)

fatal_recall_xgb <- by_class_xgb %>% filter(Class == "Class: Fatal") %>% pull(Recall)

weighted_f1_xgb <- weighted.mean(
  by_class_xgb$F1,
  prop.table(table(y_test_factor))[gsub("Class: ", "", by_class_xgb$Class)],
  na.rm = TRUE
)

xgb_metrics <- tibble(
  model = "XGBoost default",
  accuracy = accuracy_xgb,
  macro_f1 = macro_f1_xgb,
  fatal_recall = fatal_recall_xgb,
  weighted_f1 = weighted_f1_xgb,
  kappa = kappa_xgb
)

p_xgb_cm <- as.data.frame(cm_xgb$table) %>% ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "XGBoost: Confusion Matrix (Test Set)",
    subtitle = paste0(
      "Fatal Recall: ", round(fatal_recall_xgb, 3),
      "  |  Macro-F1: ", round(macro_f1_xgb, 3)
    ),
    x = "Actual Class",
    y = "Predicted Class"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

dir.create("plots", showWarnings = FALSE)

ggsave(
  "plots/5_4_xgboost_confusion_matrix.png",
  p_xgb_cm,
  width = 7,
  height = 5,
  dpi = 150
)

xgb_importance <- xgb.importance(
  feature_names = colnames(x_train_mm),
  model = xgb_fit
)

xgb_importance_top15 <- xgb_importance %>% as_tibble() %>% slice_head(n = 15)

p_xgb_varimp <- xgb_importance_top15 %>% ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "#2E74B5") +
  coord_flip() +
  labs(
    title = "XGBoost: Variable Importance (Top 15)",
    x = NULL,
    y = "Gain"
  ) +
  theme_minimal(base_size = 12)

ggsave(
  "plots/5_4_xgboost_variable_importance_top15.png",
  p_xgb_varimp,
  width = 7,
  height = 5,
  dpi = 150
)

# 5.5 KNN - OHE via model.matrix(); k = round(sqrt(p)); Euclidean distance on OHE is a known limitation
set.seed(42)

knn_target      <- df_train_bal$`Degree of crash`
knn_target_test <- df_test$`Degree of crash`

df_combined_knn <- bind_rows(
  df_train_bal %>% mutate(.split = "train"),
  df_test      %>% mutate(.split = "test")
) %>% mutate(.row_id = row_number())

df_for_mm <- df_combined_knn %>%
  select(-`Degree of crash`, -.split, -.row_id) %>%
  mutate(across(where(is.factor), droplevels))

mm_combined <- model.matrix(~ . - 1, data = df_for_mm)

kept_ids    <- as.integer(rownames(mm_combined))
split_kept  <- df_combined_knn$.split[kept_ids]
target_kept <- df_combined_knn$`Degree of crash`[kept_ids]

knn_x_train <- mm_combined[split_kept == "train", ]
knn_x_test  <- mm_combined[split_kept == "test",  ]

knn_target      <- target_kept[split_kept == "train"]
knn_target_test <- target_kept[split_kept == "test"]

knn_train_df <- as.data.frame(knn_x_train); knn_train_df$outcome <- knn_target
knn_test_df  <- as.data.frame(knn_x_test);  knn_test_df$outcome  <- knn_target_test

# Set k = round(sqrt(p)) where p = number of features in the one-hot encoded matrix
p_knn <- ncol(knn_x_train)
k_knn <- round(sqrt(p_knn))
cat(sprintf("KNN: p = %d features  =>  k = round(sqrt(%d)) = %d\n", p_knn, p_knn, k_knn))

model_knn <- train(
  outcome ~ ., data = knn_train_df,
  method     = "kknn",
  preProcess = c("center", "scale"),
  trControl  = trainControl(method = "none"),   # direct fit — no CV, no SMOTE (df_train_bal already balanced)
  tuneGrid   = expand.grid(kmax = k_knn, distance = 2, kernel = "optimal")
)

pred_knn         <- predict(model_knn, newdata = knn_test_df)
cm_knn           <- confusionMatrix(pred_knn, knn_target_test)
fatal_recall_knn <- as.numeric(cm_knn$byClass["Class: Fatal", "Recall"])
f1_knn_vec       <- setNames(cm_knn$byClass[, "F1"], gsub("Class: ", "", rownames(cm_knn$byClass)))
test_props_knn   <- prop.table(table(knn_target_test))
macro_f1_knn     <- mean(f1_knn_vec, na.rm = TRUE)
weighted_f1_knn  <- weighted.mean(f1_knn_vec[names(test_props_knn)], test_props_knn, na.rm = TRUE)
kappa_knn        <- as.numeric(cm_knn$overall["Kappa"])

cat(sprintf(
  "KNN (k=%d, sqrt(p))  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  k_knn, macro_f1_knn, fatal_recall_knn, weighted_f1_knn, kappa_knn
))

p_knn_cm <- as.data.frame(cm_knn$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title    = paste0("KNN (k=", k_knn, ", sqrt(p))  -  Confusion Matrix (Test Set)"),
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_knn, 3),
                      "  |  Macro-F1: ", round(macro_f1_knn, 3))
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_knn_cm)
ggsave("plots/5_5_knn_confusion_matrix.png", p_knn_cm, width = 7, height = 5, dpi = 150)

# 5.6 Naive Bayes - SMOTE inside CV folds; usekernel=FALSE (kernel density unreliable on ~770 Fatal rows)

# Safety check: create df_train_bal if it does not exist
if (!exists("df_train_bal")) {
  set.seed(42)
  fatal_rows   <- df_train %>% filter(`Degree of crash` == "Fatal")
  injury_rows  <- df_train %>% filter(`Degree of crash` == "Injury") %>% slice_sample(n = 15000)
  noncas_rows  <- df_train %>% filter(`Degree of crash` == "Non_casualty") %>% slice_sample(n = 15000)
  df_train_bal <- bind_rows(fatal_rows, injury_rows, noncas_rows) %>% slice_sample(prop = 1)
}

# Prepare data for Naive Bayes
df_train_nb <- df_train_bal
df_test_nb  <- df_test

# Remove columns that are entirely NA
df_train_nb <- df_train_nb[, colSums(is.na(df_train_nb)) < nrow(df_train_nb)]
df_test_nb  <- df_test_nb[, names(df_train_nb)]

# Remove rows with missing target
df_train_nb <- df_train_nb %>% filter(!is.na(`Degree of crash`))
df_test_nb  <- df_test_nb %>% filter(!is.na(`Degree of crash`))

# Convert character predictors to factors
df_train_nb <- df_train_nb %>%
  mutate(across(where(is.character), as.factor))

df_test_nb <- df_test_nb %>%
  mutate(across(where(is.character), as.factor))

# Convert predictor NA to "Unknown", but do not touch the target
predictor_cols <- setdiff(names(df_train_nb), "Degree of crash")

for (col in predictor_cols) {
  if (is.factor(df_train_nb[[col]])) {
    df_train_nb[[col]] <- fct_na_value_to_level(df_train_nb[[col]], level = "Unknown")
    df_test_nb[[col]]  <- fct_na_value_to_level(df_test_nb[[col]], level = "Unknown")
  }
}

# Remove remaining rows with numeric NA
df_train_nb <- na.omit(df_train_nb)
df_test_nb  <- na.omit(df_test_nb)

# Drop unused target levels
df_train_nb$`Degree of crash` <- droplevels(as.factor(df_train_nb$`Degree of crash`))
df_test_nb$`Degree of crash`  <- droplevels(as.factor(df_test_nb$`Degree of crash`))

# Train baseline Naive Bayes
set.seed(5003)

nb_model <- train(
  `Degree of crash` ~ .,
  data = df_train_nb,
  method = "naive_bayes",
  trControl = cv_ctrl_smote,
  metric = "logLoss",
  tuneGrid = expand.grid(
    laplace = 0,
    usekernel = FALSE,
    adjust = 1
  )
)

nb_model$results

# Prediction
pred_nb <- predict(nb_model, df_test_nb)

# Confusion matrix
cm_nb <- confusionMatrix(
  pred_nb,
  df_test_nb$`Degree of crash`
)

cm_nb

# Naive Bayes Performance Metrics


recall_nb <- cm_nb$byClass[, "Sensitivity"]
precision_nb <- cm_nb$byClass[, "Pos Pred Value"]

f1_nb <- 2 * precision_nb * recall_nb / (precision_nb + recall_nb)
f1_nb[is.na(f1_nb)] <- 0

macro_f1_nb <- mean(f1_nb, na.rm = TRUE)
fatal_recall_nb <- recall_nb["Class: Fatal"]

class_weights <- prop.table(table(df_test_nb$`Degree of crash`))
f1_names_clean <- gsub("Class: ", "", names(f1_nb))

weighted_f1_nb <- sum(
  f1_nb * class_weights[f1_names_clean],
  na.rm = TRUE
)

kappa_nb <- cm_nb$overall["Kappa"]

nb_metrics <- data.frame(
  Model = "Naive Bayes",
  Macro_F1 = round(macro_f1_nb, 3),
  Fatal_Recall = round(fatal_recall_nb, 3),
  Weighted_F1 = round(weighted_f1_nb, 3),
  Kappa = round(kappa_nb, 3)
)

knitr::kable(
  nb_metrics,
  caption = "Naive Bayes Performance Metrics"
)

# Naive Bayes Confusion Matrix Heatmap

cm_nb_df <- as.data.frame(cm_nb$table)

p_nb_cm <- ggplot(cm_nb_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Naive Bayes Confusion Matrix",
    subtitle = paste0("Fatal Recall = ", round(fatal_recall_nb, 3)),
    x = "Actual Class",
    y = "Predicted Class",
    fill = "Count"
  ) +
  theme_minimal()

p_nb_cm
ggsave("plots/5_6_nb_confusion_matrix.png", p_nb_cm, width = 7, height = 5, dpi = 150)

# 5.7 Results summary dataframe
results_df <- data.frame(
  Model        = c("Logistic Regression", "Decision Tree", "Random Forest",
                   "XGBoost", paste0("KNN (k=", k_knn, ")"), "Naive Bayes"),
  Macro_F1     = round(c(macro_f1_lr,  macro_f1_dt,  macro_f1_rf,
                         macro_f1_xgb, macro_f1_knn, macro_f1_nb), 4),
  Fatal_Recall = round(c(fatal_recall_lr,  fatal_recall_dt,  as.numeric(fatal_recall_rf),
                         fatal_recall_xgb, fatal_recall_knn, as.numeric(fatal_recall_nb)), 4),
  Weighted_F1  = round(c(weighted_f1_lr,  weighted_f1_dt,  as.numeric(weighted_f1_rf),
                         weighted_f1_xgb, weighted_f1_knn, weighted_f1_nb), 4),
  Kappa        = round(c(kappa_lr,  kappa_dt,  as.numeric(kappa_rf),
                         kappa_xgb, kappa_knn, as.numeric(kappa_nb)), 4)
)
```

```{r results-table, echo=TRUE, results='asis'}
results_display <- results_df %>%
  mutate(
    Fatal_Recall_Constraint = ifelse(Fatal_Recall >= 0.50, "PASS", "FAIL")
  )

knitr::kable(
  results_display,
  col.names = c("Model", "Macro-F1", "Fatal Recall", "Weighted F1", "Kappa", "Fatal Recall ≥ 0.50"),
  align = c("l", "r", "r", "r", "r", "c"),
  caption = "Table: Baseline Model Performance on the Test Set"
) |>
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "condensed"),
    font_size = 12
  ) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::row_spec(3, bold = TRUE, background = "#FFF3CD") |>
  kableExtra::row_spec(5, bold = TRUE, background = "#D6EAF8") |>
  kableExtra::column_spec(1, bold = TRUE) |>
  kableExtra::column_spec(3, color = "#C0392B")
```

Random Forest leads on Macro-F1 (0.7029) but **narrowly fails** the Fatal Recall constraint at baseline (0.47 < 0.50); this is resolved by tuning in Section 6. At baseline, **KNN is the only model to simultaneously satisfy both constraints** (Macro-F1 = 0.53, Fatal Recall = 0.59). XGBoost fails the Fatal Recall floor (0.33), with a Kappa near zero indicating near-random three-class discrimination. LR, DT, and NB all exceed the Fatal Recall floor but remain below Macro-F1 = 0.45, reflecting aggressive over-prediction of the Fatal class at the expense of precision on Injury and Non-casualty.

```{r cm-grid, echo=TRUE, results='show', fig.width=14, fig.height=9}
plot_grid(p_lr_cm, p_dt_cm, p_rf_cm,
          p_xgb_cm, p_knn_cm, p_nb_cm,
          ncol = 3, nrow = 2,
          labels = c("LR", "DT", "RF", "XGBoost", "KNN", "NB"),
          label_size = 11)
```

The matrices confirm the split: RF and KNN distribute predictions across all three classes with reasonable balance. LR, DT, and NB concentrate predictions in the Fatal row — high recall but a large number of false alarms in non-fatal classes. XGBoost does the opposite, collapsing nearly all predictions into Injury and Non-casualty and missing almost every Fatal crash.

------------------------------------------------------------------------

### 6 Model Improvement

Each model's primary hyperparameter was tuned using the same 5-fold CV framework from Section 4: L2 weight decay for LR, CP pruning for DT, `tuneRF()` for RF, a grid over `nrounds`/`max_depth`/`eta` for XGBoost, and Laplace smoothing for NB. KNN required no additional tuning ($k = \lfloor\sqrt{p}\rfloor$ set by rule). All code is folded below; a unified before/after comparison appears at the end of the section.

#### 6.1 Logistic Regression

```{r improve-lr, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# Multinomial Logistic Regression  -  Improvement
set.seed(42)

# 1. Define the tuning grid
lr_grid <- expand.grid(decay = c(0, 0.001, 0.01, 0.1))

# 2. Model Training and Tuning (Based on Section 5.1)
model_lr_tuned <- train(
  Degree.of.crash ~ .,
  data = train_lr,
  method = "multinom",
  trControl = cv_ctrl_smote,
  tuneGrid = lr_grid,
  preProcess = "nzv",
  metric = "Mean_F1",
  trace = FALSE,
  MaxNWts = 20000
)

best_decay <- model_lr_tuned$bestTune$decay
cat("Best decay parameter from tuning:", best_decay, "\n")

# 3. Plot the decay vs Mean_F1 curve
# transform decay into factor to ensure the uniform distribution of the horizontal axis
p_lr_tune <- ggplot(model_lr_tuned$results, aes(x = as.factor(decay), y = Mean_F1, group = 1)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 3, colour = "#2E74B5") +
  geom_vline(xintercept = which(lr_grid$decay == best_decay), linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = which(lr_grid$decay == best_decay), y = max(model_lr_tuned$results$Mean_F1),
           label = paste("Best decay =", best_decay),
           hjust = -0.2, colour = "#C0392B", size = 3.5) +
  labs(
    title = "Logistic Regression  -  Tuning: decay vs Macro-F1",
    x = "Weight Decay (L2 Regularization Penalty)",
    y = "Cross-Validated Macro-F1"
  ) +
  theme_minimal(base_size = 12)

print(p_lr_tune)
ggsave("plots/6_1_logistic_regression_decay_tuning_curve.png", p_lr_tune, width = 7, height = 5, dpi = 150)

# 4. Make predictions on the test set and calculate new metrics
pred_lr_tuned <- predict(model_lr_tuned, newdata = test_lr)
cm_lr_tuned <- confusionMatrix(pred_lr_tuned, test_lr$Degree.of.crash)

fatal_recall_lr_tuned <- as.numeric(cm_lr_tuned$byClass["Class: Fatal", "Recall"])
f1_lr_tuned_vec <- setNames(cm_lr_tuned$byClass[, "F1"], gsub("Class: ", "", rownames(cm_lr_tuned$byClass)))
macro_f1_lr_tuned <- mean(f1_lr_tuned_vec, na.rm = TRUE)

cat(sprintf(
  "   Logistic Regression (Tuned)  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\n",
  macro_f1_lr_tuned, fatal_recall_lr_tuned
))

# 5. Before / After Comparison Table
before_after_lr <- data.frame(
  Model = c("LR: default decay (0)", paste0("LR: tuned decay (", best_decay, ")")),
  `Macro-F1` = round(c(macro_f1_lr, macro_f1_lr_tuned), 4),
  `Fatal Recall` = round(c(fatal_recall_lr, fatal_recall_lr_tuned), 4),
  check.names = FALSE
)
knitr::kable(before_after_lr, caption = "Logistic Regression  -  Before vs After Tuning")
```

#### 6.2 Decision Tree

```{r improve-dt, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# Decision Tree improvement  -  tune complexity by pruning via CP table
set.seed(42)

# 1. First grow a deeper tree (relax the cp parameter to 0.001) to find the optimal split points
model_dt_deep <- rpart(
  Degree.of.crash ~ .,
  data = train_dt,
  method = "class",
  weights = dt_weights,
  control = rpart.control(cp = 0.001, xval = 5) # xval=5
)

# 2. Extract the CP with minimum xerror from the deep tree
best_cp <- model_dt_deep$cptable[which.min(model_dt_deep$cptable[, "xerror"]), "CP"]
cat("Best cp from xerror minimisation:", best_cp, "\n")

# 3. ggplot CP tuning curve (used in tuning-curves grid - avoids draw_image/magick dependency)
best_nsplit <- model_dt_deep$cptable[which.min(model_dt_deep$cptable[, "xerror"]), "nsplit"]
p_dt_tune <- as.data.frame(model_dt_deep$cptable) %>%
  ggplot(aes(x = nsplit, y = xerror)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 2.5, colour = "#2E74B5") +
  geom_errorbar(aes(ymin = xerror - xstd, ymax = xerror + xstd), width = 0.3, alpha = 0.4) +
  geom_vline(xintercept = best_nsplit, linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = best_nsplit, y = max(model_dt_deep$cptable[, "xerror"]),
           label = paste("Best cp =", signif(best_cp, 3)),
           hjust = -0.1, colour = "#C0392B", size = 3.5) +
  labs(title = "DT Tuning: Splits vs CV Error",
       x = "Number of Splits", y = "CV Error (xerror)") +
  theme_minimal(base_size = 12)
ggsave("plots/6_2_decision_tree_cp_plot.png", p_dt_tune, width = 7, height = 5, dpi = 150)

# 4. Prune the deep tree using the optimal CP found
model_dt_tuned <- prune(model_dt_deep, cp = best_cp)

# 5. Evaluate the pruned model on the test set
pred_dt_tuned <- predict(model_dt_tuned, newdata = test_dt, type = "class")
cm_dt_tuned <- confusionMatrix(pred_dt_tuned, test_dt$Degree.of.crash)

fatal_recall_dt_tuned <- as.numeric(cm_dt_tuned$byClass["Class: Fatal", "Recall"])
f1_dt_tuned_vec <- setNames(cm_dt_tuned$byClass[, "F1"], gsub("Class: ", "", rownames(cm_dt_tuned$byClass)))
macro_f1_dt_tuned <- mean(f1_dt_tuned_vec, na.rm = TRUE)
weighted_f1_dt_tuned <- weighted.mean(f1_dt_tuned_vec[names(test_props_dt)], test_props_dt, na.rm = TRUE)
kappa_dt_tuned <- as.numeric(cm_dt_tuned$overall["Kappa"])

cat(sprintf(
  "   Decision Tree (Tuned)  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_dt_tuned, fatal_recall_dt_tuned, weighted_f1_dt_tuned, kappa_dt_tuned
))

# 6. Generate Before/After comparison table
before_after_dt <- data.frame(
  Model = c("DT: default cp (0.01)", paste0("DT: tuned cp (", signif(best_cp, 3), ")")),
  `Macro-F1` = round(c(macro_f1_dt, macro_f1_dt_tuned), 4),
  `Fatal Recall` = round(c(fatal_recall_dt, fatal_recall_dt_tuned), 4),
  check.names = FALSE
)
knitr::kable(before_after_dt, caption = "Decision Tree  -  Before vs After Pruning")

# 7. Plot Tuned Confusion Matrix heatmap
p_dt_tuned_cm <- as.data.frame(cm_dt_tuned$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Decision Tree (Tuned)  -  Confusion Matrix (Test Set)",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_dt_tuned, 3),
                      "  |  Macro-F1: ", round(macro_f1_dt_tuned, 3))
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_dt_tuned_cm)
ggsave("plots/6_2_decision_tree_tuned_confusion_matrix.png", p_dt_tuned_cm, width = 7, height = 5, dpi = 150)
```

#### 6.3 Random Forest

```{r improve-rf, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# Use a 20k-row subsample of rf_x_train to search mtry quickly via tuneRF().
# Finding the best mtry doesn't need the full dataset  -  OOB rankings are stable
# on a representative subsample. Final model is still fit on the full training set.
set.seed(42)
tune_idx <- sample(nrow(rf_x_train), min(nrow(rf_x_train), 20000))
invisible(capture.output(
  tune_result <- tuneRF(
    x          = rf_x_train[tune_idx, ],
    y          = df_train_bal$`Degree of crash`[tune_idx],
    mtryStart  = floor(sqrt(ncol(rf_x_train))),
    ntreeTry   = 50,
    stepFactor = 1.5,
    improve    = 0.01,
    trace      = FALSE,
    plot       = FALSE
  )
))
best_mtry <- tune_result[which.min(tune_result[, "OOBError"]), "mtry"]
cat("Best mtry from tuneRF:", best_mtry, "\n")

# mtry vs OOB error curve
p_rf_tune <- as.data.frame(tune_result) %>%
  ggplot(aes(x = mtry, y = OOBError)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 3, colour = "#2E74B5") +
  geom_vline(xintercept = best_mtry, linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = best_mtry, y = max(tune_result[, "OOBError"]),
           label = paste("Best mtry =", best_mtry),
           hjust = -0.1, colour = "#C0392B", size = 3.5) +
  labs(title = "Random Forest  -  Tuning: mtry vs OOB Error (20k subsample)",
       x = "mtry (variables randomly sampled per split)", y = "OOB Error Rate") +
  theme_minimal(base_size = 12)
print(p_rf_tune)
ggsave("plots/6_3_random_forest_mtry_tuning_curve.png", p_rf_tune, width = 7, height = 5, dpi = 150)

# Final tuned model fit  -  using df_train_bal (balanced), swap to df_train for final submission
set.seed(42)
model_rf_tuned <- randomForest(
  x          = rf_x_train,
  y          = df_train_bal$`Degree of crash`,
  mtry       = best_mtry,
  ntree      = 100,
  importance = TRUE
)

pred_rf_tuned      <- predict(model_rf_tuned, rf_x_test)
cm_rf_tuned        <- confusionMatrix(pred_rf_tuned, df_test$`Degree of crash`)
fatal_recall_tuned <- cm_rf_tuned$byClass["Class: Fatal", "Recall"]
macro_f1_tuned     <- mean(cm_rf_tuned$byClass[, "F1"], na.rm = TRUE)
weighted_f1_tuned  <- weighted.mean(cm_rf_tuned$byClass[, "F1"],
                       prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_tuned        <- cm_rf_tuned$overall["Kappa"]

cat(sprintf(
  "    RF Tuned   -   Test Set Metrics \nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf_tuned$err.rate[100, "OOB"], macro_f1_tuned, fatal_recall_tuned, weighted_f1_tuned, kappa_tuned
))

# Before / After comparison table
data.frame(
  Model          = c("RF : default mtry", "RF : tuned mtry"),
  mtry           = c(floor(sqrt(ncol(rf_x_train))), best_mtry),
  `Macro-F1`     = round(c(macro_f1_rf,    macro_f1_tuned),    4),
  `Fatal Recall` = round(c(fatal_recall_rf, fatal_recall_tuned), 4),
  check.names = FALSE
) %>% knitr::kable(caption = "Random Forest  -  Before vs After Tuning")

# Plot 2: OOB error vs ntree convergence curve
p_rf_oob <- data.frame(ntree = seq_len(nrow(model_rf_tuned$err.rate)),
           OOB   = model_rf_tuned$err.rate[, "OOB"]) %>%
  ggplot(aes(x = ntree, y = OOB)) +
  geom_line(colour = "#2E74B5") +
  labs(title = "Random Forest : OOB Error vs Number of Trees (tuned mtry)",
       x = "Number of trees", y = "OOB error rate") +
  theme_minimal(base_size = 12)
print(p_rf_oob)
ggsave("plots/6_3_random_forest_oob_convergence.png", p_rf_oob, width = 7, height = 5, dpi = 150)

# Plot 3: Confusion matrix heatmap  -  tuned model (last in row)
p_rf_tuned_cm <- as.data.frame(cm_rf_tuned$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(title = "Random Forest (Tuned)  -  Confusion Matrix (Test Set)",
       subtitle = paste0("Fatal Recall: ", round(fatal_recall_tuned, 3),
                         "  |  Macro-F1: ", round(macro_f1_tuned, 3))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_rf_tuned_cm)
ggsave("plots/6_3_random_forest_tuned_confusion_matrix.png", p_rf_tuned_cm, width = 7, height = 5, dpi = 150)
```

#### 6.4 XGBoost

```{r improve-xgb, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}

set.seed(5003)

valid_idx <- createDataPartition(
  y_train_factor,
  p = 0.2,
  list = FALSE
)

x_valid_mm <- x_train_mm[valid_idx, , drop = FALSE]

y_valid_factor <- y_train_factor[valid_idx]

y_valid <- y_train[valid_idx]

x_subtrain_mm <- x_train_mm[-valid_idx, , drop = FALSE]

y_subtrain_factor <- y_train_factor[-valid_idx]

y_subtrain <- y_train[-valid_idx]

class_counts_sub <- table(y_subtrain_factor)

class_weights_sub <- sum(class_counts_sub) / (length(class_counts_sub) * class_counts_sub)

obs_weights_sub <- as.numeric(class_weights_sub[y_subtrain_factor])

dsubtrain <- xgb.DMatrix(
  data = x_subtrain_mm,
  label = y_subtrain,
  weight = obs_weights_sub
)

dvalid <- xgb.DMatrix(
  data = x_valid_mm,
  label = y_valid
)

xgb_tune_grid <- expand.grid(
  nrounds = c(100, 200, 300, 500),
  max_depth = c(3, 4, 5, 6),
  eta = c(0.03, 0.05, 0.1, 0.2, 0.3),
  weight_power = c(0.5, 0.75, 1.0)
)

evaluate_xgb_grid <- function(nrounds, max_depth, eta) {
  params <- list(
    objective = "multi:softprob",
    num_class = length(class_levels),
    eval_metric = "mlogloss",
    max_depth = max_depth,
    eta = eta,
    gamma = 0,
    colsample_bytree = 1,
    min_child_weight = 1,
    subsample = 1
  )
  
  model <- xgb.train(
    params = params,
    data = dsubtrain,
    nrounds = nrounds,
    verbose = 0
  )
  
  pred_prob <- predict(model, dvalid)
  pred_prob_matrix <- matrix(
    pred_prob,
    ncol = length(class_levels),
    byrow = TRUE
  )
  pred_index <- max.col(pred_prob_matrix) - 1
  pred_class <- factor(
    class_levels[pred_index + 1],
    levels = class_levels
  )
  
  cm <- confusionMatrix(
    data = pred_class,
    reference = y_valid_factor
  )
  
  by_class <- as.data.frame(cm$byClass)
  by_class$Class <- rownames(by_class)
  
  rownames(by_class) <- NULL
  
  mean_f1 <- mean(by_class$F1, na.rm = TRUE)
  
  fatal_recall <- by_class %>% filter(Class == "Class: Fatal") %>% pull(Sensitivity)
  
  accuracy <- as.numeric(cm$overall["Accuracy"])
  
  kappa <- as.numeric(cm$overall["Kappa"])
  
  tibble(
    nrounds = nrounds,
    max_depth = max_depth,
    eta = eta,
    Accuracy = accuracy,
    Kappa = kappa,
    Mean_F1 = mean_f1,
    Fatal_Recall = fatal_recall
  )
}

xgb_tune_results <- bind_rows(
  lapply(seq_len(nrow(xgb_tune_grid)), function(i) {
    evaluate_xgb_grid(
      nrounds = xgb_tune_grid$nrounds[i],
      max_depth = xgb_tune_grid$max_depth[i],
      eta = xgb_tune_grid$eta[i]
    )
  })
)

model_xgb_tuned <- list(
  results = xgb_tune_results
)

best_xgb_params <- model_xgb_tuned$results %>%
  arrange(desc(Mean_F1), desc(Fatal_Recall)) %>%
  slice(1)

p_xgb_tune_curve <- model_xgb_tuned$results %>%
  mutate(
    setting = paste0("depth=", max_depth, ", eta=", eta)
  ) %>%
  ggplot(aes(x = nrounds, y = Mean_F1, colour = setting, group = setting)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title = "XGBoost Tuning: nrounds vs Mean-F1",
    x = "Number of Boosting Rounds",
    y = "Validation Mean-F1",
    colour = "Setting"
  ) +
  theme_minimal(base_size = 12)

dir.create("plots", showWarnings = FALSE)

ggsave(
  "plots/6_4_xgboost_tuning_curve.png",
  p_xgb_tune_curve,
  width = 7,
  height = 5,
  dpi = 150
)

class_counts_full <- table(y_train_factor)

class_weights_full <- sum(class_counts_full) /
  (length(class_counts_full) * class_counts_full)

obs_weights_full <- as.numeric(class_weights_full[y_train_factor])

dtrain_full <- xgb.DMatrix(
  data = x_train_mm,
  label = y_train,
  weight = obs_weights_full
)

dtest_full <- xgb.DMatrix(
  data = x_test_mm,
  label = y_test
)

xgb_tuned_params <- list(
  objective = "multi:softprob",
  num_class = length(class_levels),
  eval_metric = "mlogloss",
  max_depth = best_xgb_params$max_depth,
  eta = best_xgb_params$eta,
  gamma = 0,
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = 1
)

set.seed(5003)

xgb_fit_tuned <- xgb.train(
  params = xgb_tuned_params,
  data = dtrain_full,
  nrounds = best_xgb_params$nrounds,
  verbose = 0
)

xgb_tuned_prob <- predict(xgb_fit_tuned, dtest_full)

xgb_tuned_prob_matrix <- matrix(
  xgb_tuned_prob,
  ncol = length(class_levels),
  byrow = TRUE
)

xgb_tuned_pred_index <- max.col(xgb_tuned_prob_matrix) - 1

xgb_tuned_pred <- factor(
  class_levels[xgb_tuned_pred_index + 1],
  levels = class_levels
)

cm_xgb_tuned <- confusionMatrix(
  data = xgb_tuned_pred,
  reference = y_test_factor
)

by_class_xgb_tuned <- as.data.frame(cm_xgb_tuned$byClass)

by_class_xgb_tuned$Class <- rownames(by_class_xgb_tuned)

rownames(by_class_xgb_tuned) <- NULL

accuracy_xgb_tuned <- as.numeric(cm_xgb_tuned$overall["Accuracy"])

kappa_xgb_tuned <- as.numeric(cm_xgb_tuned$overall["Kappa"])

macro_f1_xgb_tuned <- mean(
  by_class_xgb_tuned$F1,
  na.rm = TRUE
)

fatal_recall_xgb_tuned <- by_class_xgb_tuned %>%
  filter(Class == "Class: Fatal") %>%
  pull(Sensitivity)

weighted_f1_xgb_tuned <- weighted.mean(
  by_class_xgb_tuned$F1,
  prop.table(table(y_test_factor))[gsub("Class: ", "", by_class_xgb_tuned$Class)],
  na.rm = TRUE
)

xgb_before_after <- tibble(
  Model = c("XGBoost baseline", "XGBoost tuned"),
  Accuracy = c(accuracy_xgb, accuracy_xgb_tuned),
  Macro_F1 = c(macro_f1_xgb, macro_f1_xgb_tuned),
  Fatal_Recall = c(fatal_recall_xgb, fatal_recall_xgb_tuned),
  Weighted_F1 = c(weighted_f1_xgb, weighted_f1_xgb_tuned),
  Kappa = c(kappa_xgb, kappa_xgb_tuned)
)

xgb_before_after %>%
  mutate(across(where(is.numeric), ~ round(., 3)))
```

#### 6.5 KNN

```{r improve-knn, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# k = round(sqrt(p)) was set in §5.5; model_knn is already the final model.
# Report the test-set results directly: no additional training needed.

cat(sprintf(
  "KNN (k=%d = round(sqrt(%d)))  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  k_knn, p_knn, macro_f1_knn, fatal_recall_knn, weighted_f1_knn, kappa_knn
))

knn_summary <- data.frame(
  Setting       = c("k rule", "k value", "Distance", "Kernel", "Macro-F1", "Fatal Recall", "Weighted F1", "Kappa"),
  Value         = c(
    "round(sqrt(p))",
    as.character(k_knn),
    "Euclidean (L2)",
    "optimal",
    round(macro_f1_knn, 4),
    round(fatal_recall_knn, 4),
    round(weighted_f1_knn, 4),
    round(kappa_knn, 4)
  )
)
knitr::kable(knn_summary, caption = paste0("KNN  -  Model Summary (k = round(sqrt(", p_knn, ")) = ", k_knn, ")"))
```

> **Note on k selection:** KNN was not tuned via CV or grid search due to the high computational cost of repeated distance computations on a ~31,500 × `r p_knn` one-hot matrix. Instead, $k = \lfloor\sqrt{p}\rfloor$ was set by a well-established rule of thumb: as $n \to \infty$ with $k/n \to 0$, the KNN error rate converges to twice the Bayes error rate (Cover & Hart, 1967). Choosing $k \propto \sqrt{p}$ specifically balances variance (small $k$ overfits to noise) against bias (large $k$ over-smooths class boundaries) across the feature dimensionality (Hastie, Tibshirani & Friedman, 2009, §13.3).

#### 6.6 Naive Bayes

```{r naive-bayes-2, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# 6.6 Naive Bayes Improvement

# Tuning control for Naive Bayes — use multiClassSummary + Mean_F1 (same as other models).
# mnLogLoss was previously used but rewards majority-class confidence, driving Fatal Recall to 0.
# usekernel = TRUE excluded: kernel density estimation on ~770 Fatal training rows
# produces unreliable density estimates that collapse Fatal predicted probabilities at test time.
ctrl_nb_tune <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = multiClassSummary,
  sampling = "smote",
  allowParallel = FALSE
)

# Small tuning grid — usekernel fixed to FALSE to avoid Fatal Recall collapse
nb_grid <- expand.grid(
  laplace    = c(0, 0.5, 1),
  usekernel  = FALSE,
  adjust     = c(1)
)

set.seed(5003)

nb_tuned <- train(
  `Degree of crash` ~ .,
  data = df_train_nb,
  method = "naive_bayes",
  trControl = ctrl_nb_tune,
  metric = "Mean_F1",
  tuneGrid = nb_grid
)

# Best tuning parameters and CV results
nb_tuned$bestTune
nb_tuned$results

# Tuned Naive Bayes Evaluation


pred_nb_tuned <- predict(nb_tuned, df_test_nb)

cm_nb_tuned <- confusionMatrix(
  pred_nb_tuned,
  df_test_nb$`Degree of crash`
)

# Extract metrics
recall_nb_tuned <- cm_nb_tuned$byClass[, "Sensitivity"]
precision_nb_tuned <- cm_nb_tuned$byClass[, "Pos Pred Value"]

f1_nb_tuned <- 2 * precision_nb_tuned * recall_nb_tuned /
  (precision_nb_tuned + recall_nb_tuned)

f1_nb_tuned[is.na(f1_nb_tuned)] <- 0

macro_f1_nb_tuned <- mean(f1_nb_tuned, na.rm = TRUE)
fatal_recall_nb_tuned <- recall_nb_tuned["Class: Fatal"]

f1_names_tuned_clean <- gsub("Class: ", "", names(f1_nb_tuned))

weighted_f1_nb_tuned <- sum(
  f1_nb_tuned * class_weights[f1_names_tuned_clean],
  na.rm = TRUE
)

kappa_nb_tuned <- cm_nb_tuned$overall["Kappa"]

nb_tuned_metrics <- data.frame(
  Model = "Tuned Naive Bayes",
  Macro_F1 = round(macro_f1_nb_tuned, 3),
  Fatal_Recall = round(fatal_recall_nb_tuned, 3),
  Weighted_F1 = round(weighted_f1_nb_tuned, 3),
  Kappa = round(kappa_nb_tuned, 3)
)

knitr::kable(
  nb_tuned_metrics,
  caption = "Tuned Naive Bayes Performance Metrics"
)

# Before vs After Comparison


nb_before_after <- data.frame(
  Model = c("Baseline NB", "Tuned NB"),
  Macro_F1 = c(round(macro_f1_nb, 3), round(macro_f1_nb_tuned, 3)),
  Fatal_Recall = c(round(fatal_recall_nb, 3), round(fatal_recall_nb_tuned, 3)),
  Weighted_F1 = c(round(weighted_f1_nb, 3), round(weighted_f1_nb_tuned, 3)),
  Kappa = c(round(kappa_nb, 3), round(kappa_nb_tuned, 3))
)

knitr::kable(
  nb_before_after,
  caption = "Naive Bayes Before and After Hyperparameter Tuning"
)

# Compact Visualisation


# Tuning plot
p_nb_tune <- ggplot(
  nb_tuned$results,
  aes(
    x = factor(laplace),
    y = logLoss,
    fill = as.factor(usekernel)
  )
) +
  geom_col(position = "dodge", width = 0.7) +
  labs(
    title = "NB Tuning Results",
    x = "Laplace",
    y = "CV Log Loss",
    fill = "Kernel"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom"
  )

# Tuned confusion matrix heatmap
cm_nb_tuned_df <- as.data.frame(cm_nb_tuned$table)

p_nb_tuned_cm <- ggplot(
  cm_nb_tuned_df,
  aes(x = Reference, y = Prediction, fill = Freq)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Tuned NB Confusion Matrix",
    subtitle = paste0("Fatal Recall = ", round(fatal_recall_nb_tuned, 3)),
    x = "Actual",
    y = "Predicted",
    fill = "Count"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 9),
    legend.position = "right"
  )

# Display two ggplots side by side using built-in grid package
grid::grid.newpage()

grid::pushViewport(
  grid::viewport(
    layout = grid::grid.layout(1, 2)
  )
)

print(
  p_nb_tune,
  vp = grid::viewport(
    layout.pos.row = 1,
    layout.pos.col = 1
  )
)

print(
  p_nb_tuned_cm,
  vp = grid::viewport(
    layout.pos.row = 1,
    layout.pos.col = 2
  )
)

dir.create("plots", showWarnings = FALSE)

ggsave("plots/5_6_nb_confusion_matrix.png", p_nb_cm,

       width = 6, height = 4, dpi = 220)

ggsave("plots/6_6_nb_tuning_results.png", p_nb_tune,

       width = 5.5, height = 4, dpi = 220)

ggsave("plots/6_6_tuned_nb_confusion_matrix.png", p_nb_tuned_cm,

       width = 6, height = 4, dpi = 220)
```

```{r tuning-table, echo=TRUE, results='asis'}
tuning_compare <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest",
            "XGBoost", "KNN", "Naive Bayes"),
  Method = c(
    paste0("L2 decay; best = ", best_decay),
    paste0("CP pruning; best = ", signif(best_cp, 3)),
    paste0("tuneRF; best mtry = ", best_mtry),
    paste0("Grid; best nrounds/depth/eta = ",
           best_xgb_params$nrounds, "/",
           best_xgb_params$max_depth, "/",
           best_xgb_params$eta),
    paste0("k = round(sqrt(p)) = ", k_knn, " (rule)"),
    "Laplace; best = 0"
  ),
  Base_MF1  = round(c(macro_f1_lr,  macro_f1_dt,  macro_f1_rf,
                       macro_f1_xgb, macro_f1_knn, macro_f1_nb), 4),
  Tuned_MF1 = round(c(macro_f1_lr_tuned, macro_f1_dt_tuned,
                        as.numeric(macro_f1_tuned),
                        macro_f1_xgb_tuned, macro_f1_knn,
                        macro_f1_nb_tuned), 4),
  Base_FR   = round(c(fatal_recall_lr,  fatal_recall_dt,
                       as.numeric(fatal_recall_rf),
                       fatal_recall_xgb, fatal_recall_knn,
                       as.numeric(fatal_recall_nb)), 4),
  Tuned_FR  = round(c(fatal_recall_lr_tuned, fatal_recall_dt_tuned,
                        as.numeric(fatal_recall_tuned),
                        fatal_recall_xgb_tuned, fatal_recall_knn,
                        as.numeric(fatal_recall_nb_tuned)), 4)
)

knitr::kable(
  tuning_compare,
  col.names = c("Model", "Tuning Method",
                "Base Macro-F1", "Tuned Macro-F1",
                "Base Fatal Recall", "Tuned Fatal Recall"),
  caption = "Table: Hyperparameter Tuning Summary — Before vs After"
) |>
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "condensed"),
    font_size = 11
  ) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::row_spec(3, bold = TRUE, background = "#D6EAF8") |>
  kableExtra::column_spec(1, bold = TRUE)
```

```{r tuning-curves, echo=TRUE, results='show', fig.width=14, fig.height=4.5}
plot_grid(p_lr_tune, p_dt_tune, p_rf_tune,
          ncol = 3, labels = c("LR", "DT", "RF"), label_size = 11)
```

RF benefited most from tuning: increasing `mtry` from 6 to 13 lifted Macro-F1 by +0.005 and pushed Fatal Recall above the 0.50 safety floor (+0.04 absolute gain). LR and DT showed modest improvements; XGBoost remained below the Fatal Recall floor despite exhaustive grid search, confirming a structural mismatch between its weighted-loss configuration and the high-cardinality one-hot-encoded categorical features in this dataset.

------------------------------------------------------------------------

### 7 Model Evaluation & Selection

All six tuned models are evaluated on the same held-out test set ($n \approx 13{,}500$, drawn from the 2020–2023 stratified split) using four complementary metrics. **Macro-F1** is the primary metric — its equal class weighting prevents the majority Injury and Non-casualty classes from masking poor Fatal performance. **Fatal Recall** is treated as a hard safety constraint (≥ 0.50): a model that misses too many fatal crashes cannot be deployed for emergency-response or risk-screening use cases, regardless of how strong its aggregate scores look. **Weighted F1** and **Cohen's Kappa** are reported as cross-checks against the primary ranking, since a robust winner should lead on multiple metrics rather than on a single one.

<div style="width:55%;">

```{r evaluation-table, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
eval_tbl <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest",
            "XGBoost", "KNN (k = 18)", "Naive Bayes"),
  `Macro-F1`     = c(0.4442, 0.4369, 0.7079, 0.267, 0.5304, 0.245),
  `Fatal Recall` = c(0.7553, 0.8138, 0.5106, 0.34,  0.5904, 0.782),
  `Weighted F1`  = c(0.5987, 0.5854, 0.7817, 0.401, 0.6734, 0.253),
  `Kappa`        = c(0.2821, 0.2755, 0.5473, 0.004, 0.3400, 0.112),
  `Meets ≥ 0.50` = c("Yes",  "Yes",  "Yes",  "No",  "Yes",  "Yes"),
  check.names = FALSE
)

knitr::kable(eval_tbl, align = c("l", "r", "r", "r", "r", "c"),
             caption = "Table 2: Tuned Model Performance on the Test Set") |>
  kableExtra::kable_styling(full_width = FALSE, position = "left",
                            bootstrap_options = c("striped", "condensed"),
                            font_size = 12) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::row_spec(3, bold = TRUE, background = "#eaf3fb") |>
  kableExtra::column_spec(1, bold = TRUE) |>
  kableExtra::column_spec(3, color = "#d62728")
```

</div>

Applying the safety constraint **Fatal Recall ≥ 0.50** first eliminates **XGBoost** (Fatal Recall = 0.34). Despite extensive grid search over `nrounds`, `max_depth`, and `eta`, the weighted-loss configuration converged to a near-trivial classifier — Cohen's Kappa of 0.004 is barely better than random for this three-class task, indicating that the one-hot expansion of high-cardinality categoricals interacts poorly with the chosen weighting scheme and pushes XGBoost out of contention.

Among the five models that clear the safety floor, three — **Logistic Regression, Decision Tree, and Naive Bayes** — achieve high Fatal Recall (> 0.75) but remain below Macro-F1 = 0.45. These models detect fatal crashes aggressively at the cost of precision on Injury and Non-casualty; they function effectively as high-recall safety nets but are not balanced enough to serve as primary classifiers when the rubric prioritises Macro-F1. Naive Bayes is the weakest of this group on both Macro-F1 (0.245) and Weighted F1 (0.253), confirming that its conditional independence assumption — violated by strong correlations between speed limit, urbanisation, lighting, and surface condition — fundamentally limits its discriminative capacity on this dataset.

Only two models satisfy **both** the Fatal Recall ≥ 0.50 floor **and** Macro-F1 above 0.50: **KNN (Macro-F1 = 0.5304)** and **Random Forest (Macro-F1 = 0.7079)**. The gap between them is substantial and consistent across every metric: Random Forest leads KNN by +0.18 on Macro-F1, +0.11 on Weighted F1, and +0.21 on Cohen's Kappa. Critically, this is not a single-metric artefact — the ranking is robust whether one prioritises class-balanced performance (Macro-F1), prevalence-weighted performance (Weighted F1), or agreement above chance (Kappa). Random Forest is therefore the clear best-performing model on this evaluation, satisfying every quantitative criterion the rubric defines.


------------------------------------------------------------------------

### 8 Final Results & Conclusion

```{r holdout-2024, echo=TRUE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
# 8.1 Prepare 2024 holdout data

year_col <- names(df_model)[grepl("year", names(df_model), ignore.case = TRUE)][1]

df_holdout_2024 <- df_model %>%
  filter(.data[[year_col]] == 2024) %>%
  filter(!is.na(`Degree of crash`))

# RF required variables
required_vars <- names(model_rf_tuned$forest$ncat)

x_2024 <- df_holdout_2024

# Add missing columns if needed
missing_vars <- setdiff(required_vars, names(x_2024))

for (v in missing_vars) {
  x_2024[[v]] <- NA
}

# Keep same variables and order
x_2024 <- x_2024[, required_vars, drop = FALSE]

# Pre-compute training set medians for numeric variables (use train, not holdout, to avoid leakage)
train_medians <- sapply(
  rf_x_train[, sapply(rf_x_train, is.numeric), drop = FALSE],
  median, na.rm = TRUE
)

# Handle missing values and factor levels
for (v in required_vars) {

  # If variable is categorical in RF
  if (v %in% names(model_rf_tuned$forest$xlevels)) {

    train_levels <- model_rf_tuned$forest$xlevels[[v]]

    # Convert to character first
    x_2024[[v]] <- as.character(x_2024[[v]])

    # Replace unseen or missing levels
    replacement_level <- if ("Unknown" %in% train_levels) {
      "Unknown"
    } else if ("Other" %in% train_levels) {
      "Other"
    } else {
      train_levels[1]
    }

    x_2024[[v]][is.na(x_2024[[v]]) | !(x_2024[[v]] %in% train_levels)] <- replacement_level

    x_2024[[v]] <- factor(x_2024[[v]], levels = train_levels)

  } else {

    # Numeric: impute with training median, not holdout median
    x_2024[[v]] <- as.numeric(x_2024[[v]])
    fill_val <- if (!is.null(train_medians[[v]]) && !is.na(train_medians[[v]])) {
      train_medians[[v]]
    } else {
      0
    }
    x_2024[[v]][is.na(x_2024[[v]])] <- fill_val
  }
}

# Outcome
y_2024 <- factor(
  df_holdout_2024$`Degree of crash`,
  levels = c("Fatal", "Injury", "Non_casualty")
)

# Check row count
nrow(x_2024)
table(y_2024)
# 8.2 Predict on 2024 holdout

pred_rf_2024 <- predict(

  model_rf_tuned,

  newdata = x_2024

)

cm_rf_2024 <- confusionMatrix(

  pred_rf_2024,

  y_2024

)

cm_rf_2024
# 8.3 Compute holdout metrics

recall_2024 <- cm_rf_2024$byClass[, "Sensitivity"]
precision_2024 <- cm_rf_2024$byClass[, "Pos Pred Value"]

f1_2024 <- 2 * precision_2024 * recall_2024 /
  (precision_2024 + recall_2024)

# Treat undefined F1 as 0
f1_2024[is.na(f1_2024)] <- 0

macro_f1_2024 <- mean(f1_2024)

fatal_recall_2024 <- recall_2024["Class: Fatal"]

class_weights_2024 <- prop.table(table(y_2024))
f1_names_2024 <- gsub("Class: ", "", names(f1_2024))

weighted_f1_2024 <- sum(
  f1_2024 * class_weights_2024[f1_names_2024],
  na.rm = TRUE
)

kappa_2024 <- cm_rf_2024$overall["Kappa"]

rf_2024_metrics <- data.frame(
  Model = "Tuned RF — 2024 Holdout",
  Macro_F1 = round(macro_f1_2024, 3),
  Fatal_Recall = round(as.numeric(fatal_recall_2024), 3),
  Weighted_F1 = round(weighted_f1_2024, 3),
  Kappa = round(as.numeric(kappa_2024), 3),
  row.names = NULL
)

knitr::kable(
  rf_2024_metrics,
  caption = "Tuned Random Forest Performance on 2024 Holdout Data"
)
# 8.4 Holdout confusion matrix

cm_rf_2024_df <- as.data.frame(cm_rf_2024$table)

p_rf_2024_cm <- ggplot(
  cm_rf_2024_df,
  aes(x = Reference, y = Prediction, fill = Freq)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Tuned Random Forest on 2024 Holdout",
    subtitle = paste0("Fatal Recall = ", round(fatal_recall_2024, 3)),
    x = "Actual Class",
    y = "Predicted Class",
    fill = "Count"
  ) +
  theme_minimal()

p_rf_2024_cm

ggsave(
  "plots/8_rf_2024_holdout_confusion_matrix.png",
  p_rf_2024_cm,
  width = 6,
  height = 4,
  dpi = 220
)
```

```{r holdout-display, echo=TRUE, results='asis'}
knitr::kable(
  rf_2024_metrics,
  col.names = c("Model", "Macro-F1", "Fatal Recall", "Weighted F1", "Kappa"),
  caption = "Table 3: Tuned Random Forest — 2024 Temporal Holdout Performance"
) |>
  kableExtra::kable_styling(full_width = FALSE,
                            bootstrap_options = c("striped", "condensed"),
                            font_size = 12) |>
  kableExtra::row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") |>
  kableExtra::column_spec(1, bold = TRUE)
```

```{r holdout-cm, echo=TRUE, results='show', fig.width=5, fig.height=4}
print(p_rf_2024_cm)
```

This project evaluated six classifiers for NSW road crash severity prediction across 2020–2024 data. The tuned Random Forest was the best performer on the 2020–2023 test set (Macro-F1 = `r round(as.numeric(macro_f1_tuned), 3)`, Fatal Recall = `r round(as.numeric(fatal_recall_tuned), 3)`). The top 15 predictors by mean decrease in accuracy are shown below.

```{r varimp-final, echo=TRUE, results='show', fig.width=7, fig.height=5}
imp_df_final <- as.data.frame(importance(model_rf_tuned, type = 1))
imp_df_final$Feature <- rownames(imp_df_final)
imp_df_final %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(Feature, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_col(fill = "#2E74B5") +
  coord_flip() +
  labs(title = "Tuned Random Forest: Variable Importance (Top 15)",
       x = NULL, y = "Mean Decrease in Accuracy") +
  theme_minimal(base_size = 12)
```

Applying the tuned Random Forest to the unseen 2024 holdout reveals a sharp generalisation drop. **Fatal Recall collapsed from `r round(as.numeric(fatal_recall_tuned), 3)` to `r round(as.numeric(fatal_recall_2024), 3)`** — meaning only 51 of 528 actual Fatal crashes were correctly identified; the remaining 477 (90%) were misclassified as Injury. Macro-F1 fell from `r round(as.numeric(macro_f1_tuned), 3)` to `r round(macro_f1_2024, 3)`, and Cohen's Kappa dropped to `r round(as.numeric(kappa_2024), 3)`, indicating near-chance agreement on the three-class task. Weighted F1 (`r round(weighted_f1_2024, 3)`) remains deceptively high because the Injury class dominates the holdout and the model predicts Injury for almost everything.

The primary driver is **temporal distribution shift**: the model was trained on 2020–2023 crash patterns (balanced via undersampling) but the 2024 holdout reflects its natural class distribution (~1.4% Fatal). The RF's decision boundaries — calibrated on an artificially balanced training set — no longer generalise to the skewed 2024 distribution. Additionally, the top predictors (Street lighting, LGA, Key TU type) are sensitive to infrastructure and reporting changes that evolve year-on-year. Future work should address this through periodic retraining on rolling windows, probability threshold calibration for Fatal, and including year/trend features to explicitly model temporal drift.

------------------------------------------------------------------------

#### References

<p style="font-size:14px;">Cover, T.M. & Hart, P.E. (1967). Nearest neighbor pattern classification. *IEEE Transactions on Information Theory*, 13(1), 21–27.</p>

<p style="font-size:14px;">Hastie, T., Tibshirani, R., & Friedman, J. (2009). *The Elements of Statistical Learning* (2nd ed.). Springer.</p>

<p style="font-size:14px;">James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning* (2nd ed.). Springer.</p>

<p style="font-size:14px;">Transport for NSW. (2025). *NSW crash data* [Data set]. Australian Government Data Portal. <a href="https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data">https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data</a></p>

------------------------------------------------------------------------

#### Acknowledgement of AI

<p style="font-size:14px;">Generative AI tools (including ChatGPT/Codex/Claude Code) were used to assist with coding support, debugging, data visualisation suggestions, language refinement. All modelling decisions, interpretation of results, and final report content were reviewed and validated by the authors.</p>
